---
title: "2 - Load data and create phyloseq"
output: html_notebook
---
# Parameters and folders 
```{r}
# Phyloseq object consists of 3 parts (i.e. it is an S4 object linking different spreadsheets (S4 objects are particularly useful for handling complex data structures in bioinformatics, statistics, and other data-intensive fields) – in our case 3 (where otu_table = ARGs in each sample, sample_data = the qPCR 16S metadata, tax_table = categorisations of genes into clusters). 
seed <- 2202
set.seed(seed)

# input for ps object 
metafile <- "Input_files//VGOCOPD_metadata_completed_20210429alx_correctFormulaColumnTU.xlsx.tab"
argclusterfile <- "Input_files//ResfinderClustersFinal_Resfinder20200127_inclGeneLen+pheno_20201206b.tab"
argmappingfile <- "Input_files//02e_TotalResfinderMapped_v3.tab"
```
# Read in the raw data
```{r}
# Data were pre-processed from individual excel sheets (metadata and 'otu data') and the refined ARG clustering spreadsheet (ResfinderClustersFinal_Resfinder20200127_inclGeneLen+pheno_20201206b)
getwd()
metadata <- read.csv(metafile, header=TRUE, sep="\t", row.names=1 )
metadata

ARGclusters <- read.csv(argclusterfile, header=TRUE, sep="\t", row.names=1 )
ARGclusters

ARG.long <- read.csv(argmappingfile, header=TRUE, sep="\t")
head(ARG.long)

# Make matrix of 'otu_data' - with samples in rows and ARGs in columns
ARGcount <- ARG.long %>%
    select(SampleName, ARGname, MappedCount) %>% 
    pivot_wider(names_from=ARGname, values_from=MappedCount, values_fill=0)
ARGcount
```

# Create PS objects 

- 1: Relative 'rarefy': ARG read count data was first rarefied according to the relative correction factors per sample (phyloseq= merged.rar3)
- 2: Gene length correction: Rarefied ARG read count data was then corrected for ARG lengths (now FPK) ((read counts*1000)/ARG length)  (phyloseq= merged.rar3.fpk)
- 3: Bacterial content correction: FPK values were normalised to total bacteria (using qPCR 16-S values for each sample)
  
Since rarefaction of data before or after ARG length correction could make a limited/small (tested with artificial set), we decided to rarefy before ARG length correction.

# Non-normalised PS 
```{r}
ARGotutable <- ARGcount %>% column_to_rownames("SampleName")

# Create ps with otu table and tax table
merged <- phyloseq(otu_table(ARGotutable, taxa_are_rows=FALSE ), tax_table( as.matrix(ARGclusters)))

# somehow the apostrophes get converted from df to matrix. Therefore we import again directly the matrix into the ps object to preserve names.
merged@otu_table@.Data <- as.matrix(ARGotutable)

# Add the meta data by aligning and merging it
metadata <- metadata[match(sample_names(merged), rownames(metadata)), ]

# final PS object
merged@sam_data <- sample_data(metadata)

# Still some of the apostrophes are missing from the tax_table - copy over the correct names to the tax_table rows
rownames(merged@tax_table) <- colnames(merged@otu_table)

# make numeric value to factor variable
merged@sam_data$Enrich_pool_num <- as.factor(merged@sam_data$Enrich_pool_num)

# Save ps object 
saveRDS(merged, "Output_files//Phyloseq_objects//1_COPD_resistome_phyloseq_object.RDS")

# Save otu table
otu <- as.data.frame(otu_table(merged)) %>% rownames_to_column( "SampleID" )
write.csv(otu, paste0(psfile, ".tab" ), row.names=FALSE )

# Basic metrics of PS object
summary(sample_data(merged))
```
# Normalise PS for relative volume inputs  
```{r}
# We need to normalise/correct the relative ratios between samples.
# - Correct for relative volume inputs (using relative rarefaction)
# - Subsequently we need to correct for ARG gene length. (N.B. We tested in a small artificial example that rarefication before or after ARG length correction could make a limited/small difference (tested with artificial set...see R playground using 10 samples), we decided to do it 100% correct and rarefy before ARG length correction.
# - Normalise to bacterial content (16S qPCR data) of each sample.
# Based on the correction factors calculated from the wet-lab data we determined a correction factor by which each sample should be normalised to individually. We therefore calculated what the target read depth should be per sample based on inputs. Use that per-sample value to rarefy each sample individually.  
# how many samples are there in the phyloseq object?
nsamples <- length(sample_names(merged))

# determine which number we should rarefy to (from metadata table)
metasum <- data.frame(sample_data(merged))$RelativeCorrect_readpair_counts

#first do sample 1, then build the rest around it
i=1
m.tmp <- subset_samples(merged, sample_names(merged) == sample_names(merged)[i] )
m.tmp <- rarefy_even_depth(m.tmp, sample.size=metasum[i], trimOTUs=FALSE, rngseed=seed )
mtotal <- m.tmp

#now add remainder samples one by one after rarefy to same m4.final ps object
for( i in 2:nsamples) {
  m.tmp <- subset_samples( merged, sample_names(merged) == sample_names(merged)[i] )
  m.tmp <- rarefy_even_depth( m.tmp, sample.size=metasum[i], trimOTUs=FALSE, rngseed=seed )
  mtotal <- merge_phyloseq( mtotal, m.tmp )
}

merged.rar <- mtotal

# Inspect rarefied data
merged.rar
# Compare to non-rarefied data
rarefy_compare <- data.frame( "OriginalSum"=sample_sums(merged), "RarefiedSum"=sample_sums(merged.rar) )
rarefy_compare

## Restore taxa_names
# At some point all apostrophes got removed from the taxa names. We need to restore them, and do this here
taxnames_compare <- data.frame("org"=taxa_names(merged),"altered.rar"=taxa_names(merged.rar))
taxnames_compare

taxa_names(merged.rar) <- taxa_names(merged)
taxnames_compare2 <- data.frame("org"=taxa_names(merged),"altered.rar"=taxa_names(merged.rar))
taxnames_compare2

## Cleanup zero taxa and samples
# N.B. using the subset_taxa or subset_samples functions removes the apostrophes again from the ps objecect but prune_taxa does not.
# Filter out taxa (microbial features) in the merged.rar phyloseq object that have zero total counts across all samples. 
merged.rar2 <- prune_taxa(taxa_sums(merged.rar) > 0, merged.rar) 
merged.rar2
# Filter out samples from the merged.rar2 phyloseq object that have zero total counts. The resulting merged.rar3 object contains only samples with non-zero counts for at least one taxa.
merged.rar3 <- prune_samples(sample_sums(merged.rar2) > 0, merged.rar2)
merged.rar3

# check if taxa names still contain apostrophes
head(taxa_names(merged.rar3)) # yes names are correct

# save this ps for later use for DA analysis
saveRDS(merged.rar3, file = "Output_files//Phyloseq_objects//2_COPD_resistome_phyloseq_object_rarefied.RDS")
```
# Correct PS for ARG gene length
```{r}
# unique list of ARG and ARG length (n=369)
ARGlen <- data.frame("ARGname"=ARG.long$ARGname, "ARGlen"=ARG.long$ARGlength)
ARGlen <- unique.data.frame(ARGlen)
rownames(ARGlen) <- NULL 
ARGlen <- ARGlen %>% column_to_rownames("ARGname")
ARGlen

## ARG count to FPK
merged.rar3.fpk <- merged.rar3
# convert otu counts directly by matrix manipulation in the ps object using ARGlen df.
merged.rar3.fpk@otu_table@.Data <- merged.rar3.fpk@otu_table@.Data * 1000 / ARGlen[ col(merged.rar3.fpk@otu_table@.Data), ] 

# Check the ps
merged.rar3.fpk
head(taxa_names(merged.rar3.fpk))
head(data.frame(otu_table(merged.rar3.fpk)))

# Save partially normalised data 
saveRDS(merged.rar3.fpk, "Output_files//Phyloseq_objects//3_COPD_resistome_phyloseq_object_rarefied_FPK.RDS")

# Save otu table
otu <- as.data.frame(otu_table(merged.rar3.fpk)) %>% rownames_to_column("SampleID")
write.csv(otu, paste0(psfpkfile, ".tab"), row.names=FALSE)

# Basic metrics of normalised/ARG length-corrected PS 
summary(sample_data(merged.rar3.fpk))
merged@tax_table
```
# Phyloseq for Zenodo
```{r}
# Check if the row names match between 'merged' and 'ARGlen'
row_names_match <- all(rownames(merged@tax_table) == rownames(ARGlen))

if (row_names_match) {
  cat("Row names match between 'merged' and 'ARGlen'.\n")
} else {
  cat("Row names do not match between 'merged' and 'ARGlen'.\n")
}


# Create a new tax_table with the additional 'ARGlen' column
new_tax_table_df <- data.frame(merged@tax_table, ARGlen = ARGlen$ARGlen)
colnames <- colnames(new_tax_table_df)

new_tax_table <- tax_table(new_tax_table_df)
colnames(tax_table(new_tax_table)) <- colnames
taxa_names(new_tax_table) <- rownames(new_tax_table_df)

zenodo_ps <- merged
zenodo_ps@tax_table <- new_tax_table
zenodo_ps@tax_table


# Now keep only information we want for the zenodo ps object
# Taxonomy table
# Create a copy of the tax_table
new_tax_table_zenodo <- tax_table(zenodo_ps)
colnames(tax_table(zenodo_ps))
# Define 'tax_table_columns_to_keep' as a vector of the column names you want to keep
tax_table_columns_to_keep <- c("ARG_class", "ARGCluster", "ARGlen")
# Subset the tax_table to keep only the specified columns
new_tax_table_zenodo <- new_tax_table_zenodo[, tax_table_columns_to_keep]
# Replace the tax_table in your phyloseq object with the modified one
zenodo_ps@tax_table <- new_tax_table_zenodo

# Sample data 
new_sam_data_zenodo <- zenodo_ps@sam_data
colnames(zenodo_ps@sam_data)
sam_data_columns_to_keep <- c("copdcaco", "qPCR_16S_ngml", "Enrich_pool_num", "RelativeCorrect_readpair_counts")
new_sam_data_zenodo <- new_sam_data_zenodo[, sam_data_columns_to_keep]
zenodo_ps@sam_data <- new_sam_data_zenodo

# Otu table - keep as normal (no corrections)
zenodo_ps@otu_table

# Remove duplicate sample (13674) from the dataset
zenodo_ps <- prune_samples(sample_names(zenodo_ps) != "13674", zenodo_ps) 
zenodo_ps # Final PS object therefore now has 72 samples

resistome_samdata<- zenodo_ps@sam_data
resistome_samdata$ParticipantID <- rownames(resistome_samdata)
write.xlsx(resistome_samdata, "Output_files//Phyloseq_objects//Sample_data_resistome_VGO_COPD.xlsx", rownames= TRUE)


zenodo_ps@sam_data$copdcaco <- revalue(zenodo_ps@sam_data$copdcaco, c('1' ='COPD', '0' = 'control'))
zenodo_ps@sam_data$copdcaco
zenodo_ps
saveRDS(zenodo_ps,"Output_files//Phyloseq_objects//Phyloseq_resistome_VGO_COPD.rds")

```
# Correct PS for 16S qPCR counts
```{r}
# create a list of sample ID and associated 16S qPCR count
qPCRcounts <- data.frame("SampleID"= unique(ARG.long$SampleName), "qPCR16S"= metadata$qPCR_16S_ngml)
rownames(qPCRcounts) <- NULL
qPCRcounts
qPCRcounts <- qPCRcounts %>% column_to_rownames("SampleID")
qPCRcounts

saveRDS(qPCRcounts,"Output_files//Phyloseq_objects//qPCRcounts")

merged.rar3.fpkm <- merged.rar3.fpk
# convert otu fpk values directly by matrix manipulation in the ps object using qPCRcounts data frame 
merged.rar3.fpkm@otu_table@.Data <- merged.rar3.fpkm@otu_table@.Data / qPCRcounts [ 1:73 ,]

# check PS 
merged.rar3.fpkm
head(taxa_names(merged.rar3.fpkm))

# Manual checking of data
view(merged.rar3.fpkm@otu_table@.Data)
view(merged.rar3.fpk@otu_table@.Data)
view(merged.rar3@otu_table@.Data) 
# Manual checks of the otu table data shows me that this correction has been successful as we can see that each value has been divided by the individual ARG lengths


## Remove duplicate sample (13674) from the dataset
# There was an error in the choice and labelling of 1 control sample and we ended up with the same control sample being used for 2 case samples. It was mislabelled as '13674' hence was thought to be a different sample to '13764' which was already included in the dataset. Therefore we must remoce '13674' from the dataset. 
Final.ps <- prune_samples(sample_names(merged.rar3.fpkm) != "13674", merged.rar3.fpkm) 
Final.ps
# Final PS object therefore now has 72 samples in total (35 cases, 34 controls, 3 blanks)

# Save fully normalised & corrected data
saveRDS(Final.ps, "Output_files//Phyloseq_objects//4_COPD_resistome_phyloseq_object_rarefied_FPKM.RDS")

# Save otu table
otu_fpkm <- as.data.frame(otu_table(Final.ps)) %>% rownames_to_column("SampleID")
write.csv(otu_fpkm, paste0(psfpkmfile, ".tab" ), row.names=FALSE)
```
# Agglomerate at 90% ARG cluster level 
```{r}
# Now agglomerate the data to AMRcluster level - i.e. merge ARGs which are in the same ARGcluster (90% identity level)  
# Clustering of ARGs at the 90% identity level is classified as 'ARGCluster' in the tax data of the Final.ps object - agglomerate at this level i.e. create a new ps object with cluster 90 as default. Automatic default is the ARG level but this may not be very accurate therefore best to use cluster90 level (ARGCluster in ps object). 

# print the available taxonomic ranks
colnames(tax_table(Final.ps))
# how many ARGs are there before agglomeration in original ps object? 
Final.ps # 233 taxa (ARGs)

# Tax_glom2 function as created by AB - save this in the global environment for later use
tax_glom2 <- function(physeq, taxrank=rank_names(physeq)[1],
                     NArm=TRUE, bad_empty=c(NA, "", " ", "\t")){
  #### This part is identical to phyloseq's tax_glom
  # Error if tax_table slot is empty
  if( is.null(access(physeq, "tax_table")) ){
    stop("The tax_glom2() function requires that physeq contain a taxonomyTable")
  }
  # Error if bad taxrank
  if( !taxrank[1] %in% rank_names(physeq) ){
    stop("Bad taxrank argument. Must be among the values of rank_names(physeq)")
  }
  # Make a vector from the taxonomic data.
  CN  <- which( rank_names(physeq) %in% taxrank[1] )
  tax <- as(access(physeq, "tax_table"), "matrix")[, CN]
  # if NArm is TRUE, remove the empty, white-space, NA values from
  if( NArm ){
    keep_species <- names(tax)[ !(tax %in% bad_empty) ]
    physeq <- prune_taxa(keep_species, physeq)
  }
  # Concatenate data up to the taxrank column, use this for agglomeration
  tax <- as(access(physeq, "tax_table"), "matrix")[, 1:CN, drop=FALSE]
  tax <- apply(tax, 1, function(i){paste(i, sep=";_;", collapse=";_;")})
  #### **Speedyseq changes start here**
  ## Make the new OTU table
  # Work with taxa as rows
  if (!taxa_are_rows(physeq)) {
    physeq <- phyloseq::t(physeq)
    # Note that we need to flip back to samples as rows at the end
    needs_flip <- TRUE
  } else {
    needs_flip <- FALSE
  }
  # Starting point is a tibble with rows as taxa, to be able to combine taxa
  # with the dplyr::summarize_*() functions
  otu <- otu_table(physeq)
  tb <- otu %>%
    as("matrix") %>%
    tibble::as_tibble(rownames = "OTU")
  # We want to name each new taxon (group of merged OTUs) by its "archetype",
  # the most abundant OTU in the group
  tb <- tb %>%
    tibble::add_column(Tax = tax, Sum = taxa_sums(physeq)) %>%
    dplyr::group_by(Tax)
  # Name new taxa by the most abundant OTU; pick the first OTU in case of
  # ties (to be consistent with phyloseq)
  new_taxa_names <- tb %>%
    dplyr::top_n(1, Sum) %>%
    dplyr::slice(1) %>%
    dplyr::select(Tax, OTU)
  # Sum abundances and rename taxa
  tb0 <- tb %>%
    dplyr::summarize_at(dplyr::vars(sample_names(physeq)), sum) %>%
    dplyr::left_join(new_taxa_names, by = "Tax") %>%
    dplyr::select(OTU, dplyr::everything(), -Tax)
  # Put back into phyloseq form
  mat <- tb0 %>%
    dplyr::select(-OTU) %>%
    as("matrix")
  rownames(mat) <- tb0$OTU
  otu0 <- otu_table(mat, taxa_are_rows = TRUE)
  ## Make the new phyloseq object
  # Replacing the original otu_table with the new, smaller table will
  # automatically prune the taxonomy, tree, and refseq to the smaller set of
  # archetypal otus
  otu_table(physeq) <- otu0
  # "Empty" the taxonomy values to the right of the rank, using
  # NA_character_.
  if (CN < length(rank_names(physeq))) {
    bad_ranks <- seq(CN + 1, length(rank_names(physeq)))
    tax_table(physeq)[, bad_ranks] <- NA_character_
  }
  ## Return.
  if (needs_flip) {
    physeq <- phyloseq::t(physeq)
  }
  return(physeq)
}


# Use function tax_glom2 - 6 is the ARGCluster level in the tax_table of Final.ps object. This creates a new ps object with ARGcluster as default 
ARGcluster.aggl <- tax_glom2(Final.ps, taxrank=rank_names(Final.ps)[6], NArm=TRUE)
taxa_names(ARGcluster.aggl)
# The 'taxa_names' are not the arg cluster names, as the function just chooses one ARG name instead of selecting the cluster name.
# Check tax-table of ps object - if clustering is done correctly all other levels above 'ARG_cluster' level will be NA. 
tax_table(ARGcluster.aggl) # Yes this is the case

# How many taxa before/after agglomeration?
ntaxa(Final.ps); ntaxa(ARGcluster.aggl) # 85 ARGClusters vs 233 ARGs 
ARGcluster.aggl@tax_table[,6]
head(ARGcluster.aggl@otu_table)
taxa_names(ARGcluster.aggl)

# Function to count the number of NAs in the taxonomy table per taxonomic rank. It returns a dataframe having the NA counts and the percentage NAs of total entries per taxrank.
countTaxrankNAs <- function(  ps, filename="" )
{
  natax   <- colSums(is.na(ps@tax_table))
  tottax  <- natax + colSums(!is.na(ps@tax_table))
  perctax <- round( natax/tottax * 100, digits=2 )
  overall <- data.frame(NtaxaNAs=natax, percNAs=perctax) %>% rownames_to_column("TaxRank")
  cat( paste0("Total of ARGs = ",tottax[1],"\n"))
  if( filename > "") { write_tsv(overall, paste0(pjbase,"_TotalNAtaxaAtTaxonomicLevelsOverall.tab")) }
  return(overall)
}

countTaxrankNAs(ARGcluster.aggl) # 0 NAs in the ARG and ARG cluster columns

# Remove blanks from agglomerated PS objects
# Identify blanks and save as a new PS object
blanks <- subset_samples(ARGcluster.aggl, copdcaco == "blanc")
# Convert 'blancs' to a character vector of sample names
blank_names <- as.character(sample_names(blanks))
# Prune samples from 'ARGcluster.aggl'
ARGcluster.aggl.noblanks <- ARGcluster.aggl
ARGcluster.aggl.noblanks <- prune_samples(!sample_names(ARGcluster.aggl.noblanks)%in%blank_names, ARGcluster.aggl.noblanks)
ARGcluster.aggl.noblanks;ARGcluster.aggl # checking to ensure blanks have been removed - yes now there are 69 instead of 72 samples 

# Prune samples from 'AMRclass.aggl'
AMRclass.aggl.noblanks <- AMRclass.aggl
AMRclass.aggl.noblanks <- prune_samples(!sample_names(AMRclass.aggl.noblanks)%in%blank_names, AMRclass.aggl.noblanks)
AMRclass.aggl.noblanks;AMRclass.aggl # checking to ensure blanks have been removed - yes now there are 69 instead of 72 samples 

saveRDS(AMRclass.aggl,"Output_files//Phyloseq_objects//5_COPD_resistome_phyloseq_object_rarefied_FPKM_90clustered.RDS")
saveRDS(AMRclass.aggl.noblanks,"Output_files//Phyloseq_objects//6_COPD_resistome_phyloseq_object_rarefied_FPKM_90clustered_noblanks.RDS")
```
# Create phyloseq with livestock exposure variables
```{r}
blancs <- c("veldbl16", "veldbl3", "veldbl5")
Final.ps.noblanks <- Final.ps
Final.ps.noblanks <-  prune_samples(!sample_names(Final.ps.noblanks)%in%blancs, Final.ps.noblanks)
Final.ps.noblanks;Final.ps # checking to ensure blanks have been removed - yes now there are 69 instead of 72 samples 

# PS object which I need to add metadata to
Final.ps.noblanks
# Current meta data in this ps object
Final.ps.noblanks@sam_data

# Read in the exposure data from MdR
livestock.exp.df <- read.csv("Input_files//VGOcopdcaco_resistome_Livestpred_forBea_copdstatus.csv", header=TRUE, row.names='respnr')
# Remove the randomly added column 'X' from the dataframe 
livestock.exp.df <- subset (livestock.exp.df, select = -X)
# Remove blanks from the dataframe 
livestock.exp.df.noblanks <- livestock.exp.df[-(70:72),]
livestock.exp.df.noblanks.df <- as.data.frame(livestock.exp.df.noblanks)

# copy over original ps object 
Final.ps.noblanks.exp <- Final.ps.noblanks
sample_data(Final.ps.noblanks.exp)

sam.data.new <- as.data.frame(sample_data(livestock.exp.df.noblanks.df))
sample_data(Final.ps.noblanks.exp) <- sam.data.new
Final.ps.noblanks.exp@sam_data
```
# LUR and RF-modelled exposures
```{r}
# Import the LUR and RF estimates of E.coli, staph, tetW and mecA concentrations for each of the 69 participant
# SLR LUR-modelled E.coli, staph, tetW and mecA concentrations
LUR_RF_residential_predictions  <- read.xlsx("Input_files//LUR_RF_models//LUR_RF_residential_predictions.xlsx")

LUR_RF_residential_predictions <- data.frame(LUR_RF_residential_predictions, row.names = LUR_RF_residential_predictions$X1)
LUR_RF_residential_predictions$X1 <- NULL  


# Descriptives of participant microbial estimates
# Compute the mean for each exposure estimate
mean_values <- colMeans(LUR_RF_residential_predictions)

# Compute the standard deviation (SD) for each exposure estimate
sd_values <- apply(LUR_RF_residential_predictions, 2, sd)

# Compute the median for each exposure estimate
median_values <- apply(LUR_RF_residential_predictions, 2, median)

# Compute the 25th percentile for each exposure estimate
percentile_25_values <- apply(LUR_RF_residential_predictions, 2, quantile, probs = 0.25)

# Compute the 75th percentile for each exposure estimate
percentile_75_values <- apply(LUR_RF_residential_predictions, 2, quantile, probs = 0.75)

# Compute the minimum value for each exposure estimate 
min_values <- apply(LUR_RF_residential_predictions, 2, min)

# Compute the maximum value for each exposure estimate
max_values <- apply(LUR_RF_residential_predictions, 2, max)

# Combine the results into a new DataFrame for better visualization
summary_statistics_df <- data.frame(
  Mean = mean_values,
  SD = sd_values,
  Median = median_values,
  `25th Percentile` = percentile_25_values,
  `75th Percentile` = percentile_75_values,
  Minimum = min_values,
  Maximum = max_values
)

# Display the summary statistics
print(summary_statistics_df)
write.xlsx(summary_statistics_df, "Output_files//LUR_RF_models//summary_statistics_df_residential_modelled_exp.xlsx")

# Check distribution of modelled concentrations
hist(LUR_RF_residential_predictions$E.coli_LUR_estimates)
hist(LUR_RF_residential_predictions$Staph_LUR_estimates)
hist(LUR_RF_residential_predictions$tetW_LUR_estimates)
hist(LUR_RF_residential_predictions$mecA_LUR_estimates)

hist(LUR_RF_residential_predictions$E.coli_RF_predictions)
hist(LUR_RF_residential_predictions$Staph_RF_predictions)
hist(LUR_RF_residential_predictions$tetW_RF_predictions)
hist(LUR_RF_residential_predictions$mecA_RF_predictions)

```