---
title: "1 - Load data and create phyloseq"
author: "Beatrice Cornu Hewitt"
date: "`r Sys.time()`"
output: html_document
---

# Packages
```{r}
library(phyloseq);library(tidyr);library(dplyr);library(openxlsx); library(textshape); library(plyr)
```
# Parameters and folders 
```{r}
# Phyloseq object consists of 3 parts (i.e. it is an S4 object linking different spreadsheets (S4 objects are particularly useful for handling complex data structures in bioinformatics, statistics, and other data-intensive fields) – in our case 3 (where otu_table = ARGs in each sample, sample_data = the qPCR 16S metadata, tax_table = categorisations of genes into clusters). 
seed <- 2202
set.seed(seed)

# input for ps object 
here::here()
metafile <- here::here("Data/Raw/VGOCOPD_metadata_completed_20210429alx_correctFormulaColumnTU.xlsx.tab")
argclusterfile <- here::here("Data/Raw/ResfinderClustersFinal_Resfinder20200127_inclGeneLen+pheno_20201206b.tab")
argmappingfile <- here::here("Data/Raw/02e_TotalResfinderMapped_v3.tab")
```
# Read in the raw data
```{r}
# Data were pre-processed from individual excel sheets (metadata and 'otu data') and the refined ARG clustering spreadsheet (ResfinderClustersFinal_Resfinder20200127_inclGeneLen+pheno_20201206b)
metadata <- read.csv(metafile, header=TRUE, sep="\t", row.names=1)
ARGclusters <- read.csv(argclusterfile, header=TRUE, sep="\t", row.names=1 )
ARG.long <- read.csv(argmappingfile, header=TRUE, sep="\t")

# Make matrix of 'otu_data' - samples in rows and ARGs in columns
ARGcount <- ARG.long %>%
    select(SampleName, ARGname, MappedCount) %>% 
    pivot_wider(names_from=ARGname, values_from=MappedCount, values_fill=0)
```
# Additional metadata - Livestock exposure
```{r}
# Read in the exposure data from MdR - exposure proxies
livestock_exp1 <- read.csv("../Data/Raw/VGOcopdcaco_resistome_Livestpred_all_forBea.csv", header=TRUE, row.names='respnr')
# Remove the column 'X' from the dataframe 
livestock_exp1 <- subset(livestock_exp1, select = -X)

# Read in other livestock exposure metadata from Myrna including dispersion modelled endotoxin and PM10 
livestock_exp2 <- read.csv("../Data/Raw/VGOcopdcaco_resistome_Livestpred_forBea_copdstatus.csv", header=TRUE, row.names='respnr')
# Remove the column 'X' from the dataframe 
livestock_exp2 <- subset(livestock_exp2, select = -X)
livestock_exp2 <- subset(livestock_exp2, select = -copdcaco)

# Combine the two data frames by merging on the row names
livestock_exp <- merge(livestock_exp1, livestock_exp2, by = "row.names", all.x = TRUE, suffixes = c("", ""))
# Change column name 
names(livestock_exp)[names(livestock_exp) == "Row.names"] <- "ID"
# Remove duplicate columns if any
livestock_exp <- livestock_exp[, !duplicated(colnames(livestock_exp))]
livestock_exp_df <- livestock_exp

# Remove blanks from the dataframe
blanks <- c("veldbl16", "veldbl3", "veldbl5")
livestock_exp_df <- livestock_exp_df[!(livestock_exp_df$ID) %in% blanks, ]

# Read in the exposure modelled concentrations
# Import the LUR and RF estimates of E.coli, staph, tetW and mecA concentrations for each of the 69 participants
# SLR LUR-modelled E.coli, staph, tetW and mecA concentrations
livestock_models <- read.xlsx("../Data/Raw/LUR_RF_residential_predictions_copdstatus.xlsx")
livestock_models <- data.frame(livestock_models, row.names = livestock_models$X1)
livestock_models$X1 <- NULL

# Add to current metadata 
metadata
metadata$ID <- rownames(metadata) # Also includes blanks
# Data frame containing LUR and RF-modelled microbial concentration predictions
livestock_models
colnames(livestock_models)[colnames(livestock_models) == "pt_ID"] <- "ID"

# Join the dataframes together 
all_metadata <- merge(metadata, livestock_exp_df, by = "ID", all.x = TRUE, suffixes = c("", ""))
all_metadata <- merge(all_metadata, livestock_models, by = "ID", all.x = TRUE, suffixes = c("", ""))
# Exclude duplicate columns (copdcaco column)
all_metadata <- all_metadata[!duplicated(colnames(all_metadata))]
# Remove ID column and make ID numbers the rownames for later ps merge
rownames(all_metadata) <- all_metadata$ID
all_metadata$ID <- NULL


# only take new swabs - not lung function testing 
```

# Create PS objects 

- 1: Relative 'rarefy': ARG read count data was first rarefied according to the relative correction factors per sample (phyloseq= merged.rar3)
- 2: Gene length correction: Rarefied ARG read count data was then corrected for ARG lengths (now FPK) ((read counts*1000)/ARG length)  (phyloseq= merged.rar3.fpk)
- 3: Bacterial content correction: FPK values were normalised to total bacteria (using qPCR 16-S values for each sample)
- 4: Agglomerate to the 90% gene identity level
  
Since rarefaction of data before or after ARG length correction could make a limited/small (tested with artificial set), we decided to rarefy before ARG length correction.

# Non-normalised PS 
```{r}
ARGotutable <- ARGcount %>% column_to_rownames("SampleName")

# Create ps with otu table and tax table
merged <- phyloseq(otu_table(ARGotutable, taxa_are_rows=FALSE),
                   tax_table(as.matrix(ARGclusters)))

# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 369 taxa and 73 samples ]
# tax_table()   Taxonomy Table:    [ 369 taxa by 11 taxonomic ranks ]

# somehow the apostrophes get converted from df to matrix. Therefore we import again directly the matrix into the ps object to preserve names.
merged@otu_table@.Data <- as.matrix(ARGotutable)

# Align and add metadata to the ps object
all_metadata <- all_metadata[match(sample_names(merged), rownames(all_metadata)), ]
merged@sam_data <- sample_data(all_metadata)

# Some of the apostrophes are missing from the tax_table rownames
# Copy over the correct names from the otu table
rownames(merged@tax_table) <- colnames(merged@otu_table)
# Make enrichment pool number to a factor variable
merged@sam_data$Enrich_pool_num <- as.factor(merged@sam_data$Enrich_pool_num)

# Save ps object 
saveRDS(merged, "../Output_files/Phyloseq_objects/1_COPD_resistome_phyloseq_object.RDS")
```
# Normalise PS for relative volume inputs  
```{r}
# We need to normalise/correct the relative ratios between samples.
# - Correct for relative volume inputs (using relative rarefaction)
# - Subsequently we need to correct for ARG gene length. (N.B. We tested in a small artificial example that rarefication before or after ARG length correction could make a limited/small difference (tested with artificial set...see R playground using 10 samples), we decided to do it 100% correct and rarefy before ARG length correction.
# - Normalise to bacterial content (16S qPCR data) of each sample.
# Based on the correction factors calculated from the wet-lab data we determined a correction factor by which each sample should be normalised to individually. We therefore calculated what the target read depth should be per sample based on inputs. Use that per-sample value to rarefy each sample individually.  
# i.e. The rarefaction depth is determined based on the relative corrected read pair counts provided in the metadata.
# how many samples are there in the phyloseq object?
nsamples <- length(sample_names(merged))

# determine which number we should rarefy to (from metadata table)
metasum <- data.frame(sample_data(merged))$RelativeCorrect_readpair_counts

#first do sample 1, then build the rest around it
i=1
m.tmp <- subset_samples(merged, sample_names(merged) == sample_names(merged)[i] )
m.tmp <- rarefy_even_depth(m.tmp, sample.size=metasum[i], trimOTUs=FALSE, rngseed=seed )
mtotal <- m.tmp

#now add remainder samples one by one after rarefy to same m4.final ps object
for( i in 2:nsamples) {
  m.tmp <- subset_samples( merged, sample_names(merged) == sample_names(merged)[i] )
  m.tmp <- rarefy_even_depth( m.tmp, sample.size=metasum[i], trimOTUs=FALSE, rngseed=seed )
  mtotal <- merge_phyloseq( mtotal, m.tmp )
}
# set.seed(2202) was used to initialize repeatable random subsampling.
merged.rar <- mtotal
merged.rar
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 369 taxa and 73 samples ]
# sample_data() Sample Data:       [ 73 samples by 20 sample variables ]
# tax_table()   Taxonomy Table:    [ 369 taxa by 11 taxonomic ranks ]

# Compare to non-rarefied data
rarefy_compare <- data.frame( "OriginalSum"=sample_sums(merged), "RarefiedSum"=sample_sums(merged.rar) )
rarefy_compare
merged; merged.rar

# All apostrophes got removed from the taxa names, restore these in the new ps 
print(data.frame("original"=taxa_names(merged),"rarefied"=taxa_names(merged.rar)))
# Restore taxa_names
taxa_names(merged.rar) <- taxa_names(merged)
print(data.frame("original"=taxa_names(merged),"rarefied"=taxa_names(merged.rar)))# names restored

# Filter out ARGs in the merged.rar phyloseq object with zero total counts 
merged.rar2 <- prune_taxa(taxa_sums(merged.rar) > 0, merged.rar) 
merged.rar2
# Filter out samples from the merged.rar2 phyloseq object that have zero total counts
merged.rar3 <- prune_samples(sample_sums(merged.rar2) > 0, merged.rar2)
merged.rar3
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 233 taxa and 73 samples ]
# sample_data() Sample Data:       [ 73 samples by 20 sample variables ]
# tax_table()   Taxonomy Table:    [ 233 taxa by 11 taxonomic ranks ]

# save this ps for later use for DA analysis
saveRDS(merged.rar3, file="../Output_files/Phyloseq_objects/2_COPD_resistome_phyloseq_object_rarefied.RDS")
```
# Correct PS for ARG gene length
```{r}
# unique list of ARG and ARG length (n=369)
ARGlen <- data.frame("ARGname"=ARG.long$ARGname, "ARGlen"=ARG.long$ARGlength)
ARGlen <- unique.data.frame(ARGlen) # Remove duplicate rows to keep only unique ARGs
rownames(ARGlen) <- NULL 
ARGlen <- ARGlen %>% column_to_rownames("ARGname")
ARGlen

# ARGlen$ARGname <- rownames(ARGlen)
write.xlsx(ARGlen, "../Output_files/Phyloseq_objects/ARGlengths.xlsx")

# ARG count to FPK
merged.rar3.fpk <- merged.rar3
# Convert otu counts directly by matrix manipulation in the ps object using ARGlen df.
# arg_length_vector <- setNames(ARGlen$ARGlen, ARGlen$ARGname)
# # Use the named vector for indexing
# merged.rar3.fpk@otu_table@.Data <- merged.rar3.fpk@otu_table@.Data * 1000 / arg_length_vector[colnames(merged.rar3.fpk@otu_table@.Data)]

merged.rar3.fpk@otu_table@.Data <- merged.rar3.fpk@otu_table@.Data * 1000 / ARGlen[ col(merged.rar3.fpk@otu_table@.Data), ] 
head(merged.rar3.fpk@otu_table@.Data) # Check conversion is correct 

# Check the ps to make sure that the correction was done correctly
head(taxa_names(merged.rar3.fpk))
head(data.frame(otu_table(merged.rar3.fpk)))
head(data.frame(otu_table(merged.rar3)))
head(ARGlen) # Yes the correction is correct - scaled by a factor of 1000, then divided by their corresponding ARG length

# Save partially normalised data 
saveRDS(merged.rar3.fpk, "../Output_files/Phyloseq_objects/3_COPD_resistome_phyloseq_object_rarefied_FPK.rds")

# Basic metrics of normalised/ARG length-corrected PS 
summary(sample_data(merged.rar3.fpk))
```
# Correct PS for 16S qPCR counts
```{r}
# create a list of sample ID and associated 16S qPCR count
qPCRcounts <- data.frame("SampleID"= unique(ARG.long$SampleName), "qPCR16S"= metadata$qPCR_16S_ngml)
rownames(qPCRcounts) <- NULL
qPCRcounts <- qPCRcounts %>% column_to_rownames("SampleID")
qPCRcounts

saveRDS(qPCRcounts,"../Output_files/Phyloseq_objects/qPCRcounts")

merged.rar3.fpkm <- merged.rar3.fpk

# convert otu fpk values directly by matrix manipulation 
merged.rar3.fpkm@otu_table@.Data <- merged.rar3.fpkm@otu_table@.Data / qPCRcounts [ 1:73 ,]
head(taxa_names(merged.rar3.fpkm))

# Manual checking of data
head(merged.rar3.fpkm@otu_table@.Data)
head(merged.rar3.fpk@otu_table@.Data)
head(merged.rar3@otu_table@.Data)
head(qPCRcounts)
# Manual checks of the otu table shows that this correction has been successful

# Remove duplicate sample (13674) from the dataset
# There was an error in the choice and labelling of 1 control sample and we ended up with the same control sample being used for 2 case samples. It was mislabelled as '13674' hence was thought to be a different sample to '13764' which was already included in the dataset. Therefore we must remoce '13674' from the dataset. 
ps <- prune_samples(sample_names(merged.rar3.fpkm) != "13674", merged.rar3.fpkm) 
ps
# This ps object therefore now has 72 samples in total (35 cases, 34 controls, 3 blanks)
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 233 taxa and 72 samples ]
# sample_data() Sample Data:       [ 72 samples by 20 sample variables ]
# tax_table()   Taxonomy Table:    [ 233 taxa by 11 taxonomic ranks ]

# Save fully normalised & corrected data
saveRDS(ps, "../Output_files/Phyloseq_objects/4_COPD_resistome_phyloseq_object_rarefied_FPKM.RDS")

# Remove blanks from ps  
ps_noblanks <- ps
blanks <- c("veldbl16", "veldbl3", "veldbl5")
ps_noblanks <- prune_samples(!sample_names(ps_noblanks)%in%blanks, ps_noblanks)
ps; ps_noblanks # now 69 true samples, 3 blanks removed

saveRDS(ps, "../Output_files/Phyloseq_objects/4b_COPD_resistome_phyloseq_object_rarefied_FPKM_noblanks.RDS")


# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 233 taxa and 72 samples ]
# sample_data() Sample Data:       [ 72 samples by 237 sample variables ]
# tax_table()   Taxonomy Table:    [ 233 taxa by 11 taxonomic ranks ]
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 233 taxa and 69 samples ]
# sample_data() Sample Data:       [ 69 samples by 237 sample variables ]
# tax_table()   Taxonomy Table:    [ 233 taxa by 11 taxonomic ranks ]
```
# Agglomerate at 90% ARG cluster level 
```{r}
# Now agglomerate the data to ARGcluster level - i.e. merge ARGs which are in the same ARGcluster (90% identity level)  
# Clustering of ARGs at the 90% identity level is classified as 'ARGCluster' in the taxonomy table of the Final.ps object
ps # 233 taxa (ARGs) pre agglomeration

# Tax_glom2 function as created by AB - save this in the global environment for later use
tax_glom2 <- function(physeq, taxrank=rank_names(physeq)[1],
                     NArm=TRUE, bad_empty=c(NA, "", " ", "\t")){
  #### This part is identical to phyloseq's tax_glom
  # Error if tax_table slot is empty
  if( is.null(access(physeq, "tax_table")) ){
    stop("The tax_glom2() function requires that physeq contain a taxonomyTable")
  }
  # Error if bad taxrank
  if( !taxrank[1] %in% rank_names(physeq) ){
    stop("Bad taxrank argument. Must be among the values of rank_names(physeq)")
  }
  # Make a vector from the taxonomic data.
  CN  <- which( rank_names(physeq) %in% taxrank[1] )
  tax <- as(access(physeq, "tax_table"), "matrix")[, CN]
  # if NArm is TRUE, remove the empty, white-space, NA values from
  if( NArm ){
    keep_species <- names(tax)[ !(tax %in% bad_empty) ]
    physeq <- prune_taxa(keep_species, physeq)
  }
  # Concatenate data up to the taxrank column, use this for agglomeration
  tax <- as(access(physeq, "tax_table"), "matrix")[, 1:CN, drop=FALSE]
  tax <- apply(tax, 1, function(i){paste(i, sep=";_;", collapse=";_;")})
  #### **Speedyseq changes start here**
  ## Make the new OTU table
  # Work with taxa as rows
  if (!taxa_are_rows(physeq)) {
    physeq <- phyloseq::t(physeq)
    # Note that we need to flip back to samples as rows at the end
    needs_flip <- TRUE
  } else {
    needs_flip <- FALSE
  }
  # Starting point is a tibble with rows as taxa, to be able to combine taxa
  # with the dplyr::summarize_*() functions
  otu <- otu_table(physeq)
  tb <- otu %>%
    as("matrix") %>%
    tibble::as_tibble(rownames = "OTU")
  # We want to name each new taxon (group of merged OTUs) by its "archetype",
  # the most abundant OTU in the group
  tb <- tb %>%
    tibble::add_column(Tax = tax, Sum = taxa_sums(physeq)) %>%
    dplyr::group_by(Tax)
  # Name new taxa by the most abundant OTU; pick the first OTU in case of
  # ties (to be consistent with phyloseq)
  new_taxa_names <- tb %>%
    dplyr::top_n(1, Sum) %>%
    dplyr::slice(1) %>%
    dplyr::select(Tax, OTU)
  # Sum abundances and rename taxa
  tb0 <- tb %>%
    dplyr::summarize_at(dplyr::vars(sample_names(physeq)), sum) %>%
    dplyr::left_join(new_taxa_names, by = "Tax") %>%
    dplyr::select(OTU, dplyr::everything(), -Tax)
  # Put back into phyloseq form
  mat <- tb0 %>%
    dplyr::select(-OTU) %>%
    as("matrix")
  rownames(mat) <- tb0$OTU
  otu0 <- otu_table(mat, taxa_are_rows = TRUE)
  ## Make the new phyloseq object
  # Replacing the original otu_table with the new, smaller table will
  # automatically prune the taxonomy, tree, and refseq to the smaller set of
  # archetypal otus
  otu_table(physeq) <- otu0
  # "Empty" the taxonomy values to the right of the rank, using
  # NA_character_.
  if (CN < length(rank_names(physeq))) {
    bad_ranks <- seq(CN + 1, length(rank_names(physeq)))
    tax_table(physeq)[, bad_ranks] <- NA_character_
  }
  ## Return.
  if (needs_flip) {
    physeq <- phyloseq::t(physeq)
  }
  return(physeq)
}

# Use function tax_glom2 - 6 is the ARGCluster level in the tax_table of Final.ps object. This creates a new ps object with ARGcluster as default 
ps_aggl <- ps
ps_aggl <- tax_glom2(ps_aggl, taxrank=rank_names(ps_aggl)[6], NArm=TRUE)
rank_names(ps_aggl)[6] # ARGCluster taxonomic rank
taxa_names(ps_aggl)
# The 'taxa_names' are not the arg cluster names, as the function just chooses one ARG name instead of selecting the real cluster name.
# Check tax table of ps object - if clustering is done correctly all other levels above 'ARG_cluster' level will be NA. 
tax_table(ps_aggl) # Yes this is the case

# How many taxa before/after agglomeration?
ntaxa(ps); ntaxa(ps_aggl) # 85 ARGs now (ARGCluster 90 level) vs 233 ARGs 
ps_aggl@tax_table[,6] # inspect cluster names
taxa_names(ps_aggl)

# Remove blanks from ps_aggl
ps_aggl_noblanks <- ps_aggl
blanks <- c("veldbl16", "veldbl3", "veldbl5")
ps_aggl_noblanks <- prune_samples(!sample_names(ps_aggl_noblanks)%in%blanks, ps_aggl_noblanks)
ps_aggl_noblanks;ps_aggl # blanks removed - 69 samples remaining 

# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 85 taxa and 69 samples ]
# sample_data() Sample Data:       [ 69 samples by 237 sample variables ]
# tax_table()   Taxonomy Table:    [ 85 taxa by 11 taxonomic ranks ]
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 85 taxa and 72 samples ]
# sample_data() Sample Data:       [ 72 samples by 237 sample variables ]
# tax_table()   Taxonomy Table:    [ 85 taxa by 11 taxonomic ranks ]

saveRDS(ps_aggl,"../Output_files/Phyloseq_objects/5_COPD_resistome_phyloseq_object_rarefied_FPKM_90clustered.RDS")
saveRDS(ps_aggl_noblanks,"../Output_files/Phyloseq_objects/6_COPD_resistome_phyloseq_object_rarefied_FPKM_90clustered_noblanks.RDS")
```
# Summary of all phyloseq objects
```{r}
# These are the corrections made on the data: 
# - 1: Relative 'rarefy': ARG read count data was first rarefied according to the relative correction factors per sample (phyloseq= merged.rar3)
# - 2: Gene length correction: Rarefied ARG read count data was then corrected for ARG lengths (now FPK) ((read counts*1000)/ARG length)  (phyloseq= merged.rar3.fpk)
# - 3: Bacterial content correction: FPK values were normalised to total bacteria (using qPCR 16-S values for each sample)
# - 4: Agglomerate to the 90% gene identity level

ps # ps which has undergone steps 1-3 (not agglomerated), all samples
ps_noblanks  # ps which has undergone steps 1-3 (not agglomerated), 3 blanks removed

ps_aggl # ps which has undergone all steps 1-4 
ps_aggl_noblanks # ps which has undergone all steps 1-4, 3 blanks removed

# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 233 taxa and 72 samples ]
# sample_data() Sample Data:       [ 72 samples by 237 sample variables ]
# tax_table()   Taxonomy Table:    [ 233 taxa by 11 taxonomic ranks ]
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 233 taxa and 69 samples ]
# sample_data() Sample Data:       [ 69 samples by 237 sample variables ]
# tax_table()   Taxonomy Table:    [ 233 taxa by 11 taxonomic ranks ]
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 85 taxa and 72 samples ]
# sample_data() Sample Data:       [ 72 samples by 237 sample variables ]
# tax_table()   Taxonomy Table:    [ 85 taxa by 11 taxonomic ranks ]
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 85 taxa and 69 samples ]
# sample_data() Sample Data:       [ 69 samples by 237 sample variables ]
# tax_table()   Taxonomy Table:    [ 85 taxa by 11 taxonomic ranks ]
```
# Summary statistics of livestock exposure
```{r}
# Descriptive statistics of participant microbial estimates
mean_values <- colMeans(livestock_models[,3:10])
# Compute the standard deviation (SD) for each exposure estimate
sd_values <- apply(livestock_models, 2, sd)
# Compute the median for each exposure estimate
median_values <- apply(livestock_models, 2, median)
# Compute the 25th percentile for each exposure estimate
percentile_25_values <- apply(livestock_models, 2, quantile, probs = 0.25)
# Compute the 75th percentile for each exposure estimate
percentile_75_values <- apply(livestock_models, 2, quantile, probs = 0.75)
# Compute the minimum value for each exposure estimate 
min_values <- apply(livestock_models, 2, min)
# Compute the maximum value for each exposure estimate
max_values <- apply(livestock_models, 2, max)
# Combine the results into a new DataFrame for better visualization
summary_statistics_df <- data.frame(
  Mean = mean_values,
  SD = sd_values,
  Median = median_values,
  `25th Percentile` = percentile_25_values,
  `75th Percentile` = percentile_75_values,
  Minimum = min_values,
  Maximum = max_values
)

# Display the summary statistics
print(summary_statistics_df)
write.xlsx(summary_statistics_df, "Output_files/LUR_RF_models/summary_statistics_df_residential_modelled_exp.xlsx")

# Check distribution of modelled concentrations
hist(livestock_models$E.coli_LUR_estimates)
hist(livestock_models$Staph_LUR_estimates)
hist(livestock_models$tetW_LUR_estimates)
hist(livestock_models$mecA_LUR_estimates)

hist(livestock_models$E.coli_RF_predictions)
hist(livestock_models$Staph_RF_predictions)
hist(livestock_models$tetW_RF_predictions)
hist(livestock_models$mecA_RF_predictions)
```
# Study population characteristics
```{r}
# Microbiome ps from Warner contains metadata for the participants (this is the same for the resistome data as collected at the same period)
microbiome.ps <- readRDS("../Data/ps_COPD_d2_f_01%in2samples_08APR2021")
microbiome.ps
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 1092 taxa and 356 samples ]
# sample_data() Sample Data:       [ 356 samples by 47 sample variables ]
# tax_table()   Taxonomy Table:    [ 1092 taxa by 7 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 1092 reference sequences ]

# This ps metadata is missing data for participant 19586
# We need to extract this data from VGO2 data
VGO2 <- read.csv("../Data/Raw/Complete dataset 2014 VGO3 CP 20220618.csv")
# We also want to present livestock info for each participant - extract information on all farms within 3000m for each participant
ps_aggl_noblanks@sam_data$AllFarm.3000m

# Create metadata df from microbiome ps
metadata_microbiome <- data.frame(sample_data(microbiome.ps))
# Prune microbiome data to contain only time point 1 (T1) - all participants have sample taken at T1
vector.OP.time <- "T1"
microbiome.ps.T1 <- prune_samples(microbiome.ps@sam_data$OP_time %in% vector.OP.time, microbiome.ps)
metadata_microbiome <- data.frame(sample_data(microbiome.ps.T1))
metadata_microbiome$SampleID <- rownames(metadata_microbiome)
metadata_microbiome$SampleID <- sub("T1$", "", metadata_microbiome$SampleID)
rownames(metadata_microbiome) <- NULL
# Create a vector of sample IDs and then prune the microbiome ps to just include these samples 
resistome_sampleIDs <- c('10279','10839','11143','11502','11588','11814','12927','12965','12980','13571','13743','13764','13908','14587','15035','15391','15446','15573','16236','16716','16994','17053','17918','18439','18859','19586','20148','20264','20278','20407','21130','21167','21397','21548','21696','27187','27446','27641','27984','28038','29117','29401','29593','29745','30113','30148','30256','30491','30972','31465','31516','31563','31637','32372','32530','33076','33579','35287','35537','36022','36166','36424','36503','36845','37083','37250','38249','38407','38776')
# Extract rows from metadata based on resistome_sampleIDs
metadata_filtered <- filter(metadata_microbiome, SampleID %in% resistome_sampleIDs)
selected_columns <- c("SampleID", "gender", "age", "bmi", "COPDgold", "farmchild", "smokeever", "packyears", "educationlevel")
# Use select to filter columns in metadata_filtered
metadata_filtered <- select(metadata_filtered, all_of(selected_columns))

# Manually add data for participant 19586 from the VGO2 data
row_19586 <- subset(VGO2, respnr == 19586)
colnames(row_19586)[colnames(row_19586) == "respnr"] <- "SampleID"
colnames(row_19586)[colnames(row_19586) == "copd_gold"] <- "COPDgold"
row_19586_filtered <- select(row_19586, all_of(selected_columns))

# Ensure columns in both datasets are the same data type
row_19586_filtered$SampleID <- as.character(row_19586_filtered$SampleID)
row_19586_filtered$gender <- as.character(row_19586_filtered$gender)
row_19586_filtered$COPDgold <- as.factor(row_19586_filtered$COPDgold)
row_19586_filtered$farmchild <- as.character(row_19586_filtered$farmchild)
row_19586_filtered$smokeever <- as.factor(row_19586_filtered$smokeever)
row_19586_filtered$educationlevel <- as.numeric(row_19586_filtered$educationlevel)

metadata_tab1 <- bind_rows(metadata_filtered, row_19586_filtered)
# VGO2 data had different "smokeever" coding where 0=never and 1= former smoker, change this for participant 19586
# Find the row index where SampleID is 19586
row_index <- which(metadata_tab1$SampleID == "19586")
# Change the value for "smokeever" in the specific row
metadata_tab1[row_index, "smokeever"] <- "former smoker"
# Now, the df merged has the metadata for all resistome participants including the one missing from microbiome dataset.

# Now add some livestock info
ps_aggl_noblanks@sam_data$SampleID <- rownames(ps_aggl_noblanks@sam_data)
farm_info <- data.frame(SampleID = ps_aggl_noblanks@sam_data$SampleID, ps_aggl_noblanks@sam_data$AllFarm.3000m)
rownames(farm_info) <- NULL
# Merge the datasets on SampleID
merged_exp <- merge(metadata_tab1, farm_info, by = "SampleID", all.x = TRUE)
colnames(merged_exp)[colnames(merged_exp) == "Final.ps.noblanks.exp.sam_data.AllFarm.3000m"] <- "AllFarm.3000m"
# Create copdcaco column in data set
merged_exp$copdcaco <- ifelse(merged_exp$COPDgold == 0, "Control", "COPD")

# Convert blank cells to 'NA'
merged_exp <- merged_exp %>%
  mutate_all(as.character) %>%  # Convert all columns to character
  mutate_all(~ na_if(., ""))    # Replace blank cells with NA

# Ensure variable data types are correct 
merged_exp$gender <- as.factor(merged_exp$gender)
merged_exp$COPDgold <- as.factor(merged_exp$COPDgold)
merged_exp$farmchild <- as.factor(merged_exp$farmchild)
merged_exp$smokeever <- as.factor(merged_exp$smokeever)
merged_exp$educationlevel <- as.factor(merged_exp$educationlevel)
merged_exp$age <- as.numeric(merged_exp$age)
merged_exp$bmi <- as.numeric(merged_exp$bmi)
merged_exp$packyears <- as.numeric(merged_exp$packyears)
merged_exp$AllFarm.3000m <- as.numeric(merged_exp$AllFarm.3000m)

# Function to compute p-values between COPD case and control groups
pvalue <- function(x, ...) {
    # Construct vectors of data y, and groups (strata) g
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
        # For numeric variables, perform a standard 2-sample t-test
        p <- t.test(y ~ g)$p.value
    } else {
        # For categorical variables, perform a chi-squared test of independence
        p <- chisq.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}
# Numeric variables (age, age, bmi, packyears, AllFarm.3000m): ANOVA performed. The p-value is derived from the ANOVA results.
# Categorical Variables (both binary and with > 2 levels)(gender, COPDgold, farmchild, smokeever, educationlevel): chi-squared test is performed.

# Relabel populations 
merged_exp$copdcaco <- 
  factor(merged_exp$copdcaco, 
         levels=c('Control','COPD'),
         labels=c("Control population",
                  "COPD patients"))

merged_exp <- merged_exp %>%
  mutate(gender = recode(gender, `0` = "Male", `1` = "Female"))%>%
  mutate(farmchild = recode(farmchild, `0` = "No", `1` = "Yes"))%>%
  mutate(educationlevel= recode(farmchild, `0` = "Low", `1` = "Medium", `2` = "High"))%>%
  mutate(smokeever= recode(smokeever, `0` = "Never", `1` = "Former smoker"))

# Set labels for table 
label(merged_exp$gender) <- "Gender"
label(merged_exp$age) <- "Age"
label(merged_exp$bmi) <- "BMI"
label(merged_exp$COPDgold) <- "COPD GOLD grade"
label(merged_exp$farmchild) <- "Childhood on farm"
label(merged_exp$smokeever) <- "Smoking status"
label(merged_exp$packyears) <- "Pack-years of cigarettes smoked"
label(merged_exp$educationlevel) <- "Education level"
label(merged_exp$AllFarm.3000m) <- "Number of farms within 3km radius"

# Use table1 to create table 
resistome_table1 <- table1(
  ~ gender + age + bmi + COPDgold + farmchild + smokeever + packyears + educationlevel + AllFarm.3000m| copdcaco,
  data = merged_exp, 
  extra.col = list(`P-value`=pvalue),
  overall = F
)
# NB merged_exp for one CP is missing - the one from which microbiome data was removed
getwd()
write.csv(resistome_table1, file = "../Output_files/Descriptives/table1.csv")
write.xlsx(resistome_table1, file ="../Output_files/Descriptives/table1.xlsx" )

# Age and bmi p-values are the same - double check this 
ttest_result_age <- t.test(age ~ copdcaco, data = merged_exp)
# Print the result
print(ttest_result_age)
hist(merged_exp$age)

ttest_result_bmi <- t.test(bmi ~ copdcaco, data = merged_exp)
# Print the result
print(ttest_result_bmi)

hist(merged_exp$age)
hist(merged_exp$bmi)
```

# Phyloseq for Zenodo
```{r}
# Check if the row names match between 'merged' and 'ARGlen'
rownames(merged@tax_table)
row_names_match <- all(rownames(merged@tax_table) == rownames(ARGlen))

if (row_names_match) {
  cat("Row names match between 'merged' and 'ARGlen'.\n")
} else {
  cat("Row names do not match between 'merged' and 'ARGlen'.\n")
}

# Create a new tax_table with the additional 'ARGlen' column
new_tax_table_df <- data.frame(merged@tax_table, ARGlen = ARGlen$ARGlen)
colnames <- colnames(new_tax_table_df)

new_tax_table <- tax_table(new_tax_table_df)
colnames(tax_table(new_tax_table)) <- colnames
taxa_names(new_tax_table) <- rownames(new_tax_table_df)

zenodo_ps <- merged
zenodo_ps@tax_table <- new_tax_table
zenodo_ps@tax_table

# Now keep only information we want for the zenodo ps object
# Taxonomy table
# Create a copy of the tax_table
new_tax_table_zenodo <- tax_table(zenodo_ps)
colnames(tax_table(zenodo_ps))
# Define 'tax_table_columns_to_keep' as a vector of the column names you want to keep
tax_table_columns_to_keep <- c("ARG_class", "ARGCluster", "ARGlen")
# Subset the tax_table to keep only the specified columns
new_tax_table_zenodo <- new_tax_table_zenodo[, tax_table_columns_to_keep]
# Replace the tax_table in your phyloseq object with the modified one
zenodo_ps@tax_table <- new_tax_table_zenodo

# Sample data 
new_sam_data_zenodo <- zenodo_ps@sam_data
colnames(zenodo_ps@sam_data)
sam_data_columns_to_keep <- c("copdcaco", "qPCR_16S_ngml", "Enrich_pool_num", "RelativeCorrect_readpair_counts")
new_sam_data_zenodo <- new_sam_data_zenodo[, sam_data_columns_to_keep]
zenodo_ps@sam_data <- new_sam_data_zenodo

# Otu table - keep as normal (no corrections)
zenodo_ps@otu_table

# Remove duplicate sample (13674) from the dataset
zenodo_ps <- prune_samples(sample_names(zenodo_ps) != "13674", zenodo_ps) 
zenodo_ps # Final PS object therefore now has 72 samples

resistome_samdata<- zenodo_ps@sam_data
resistome_samdata$ParticipantID <- rownames(resistome_samdata)
write.xlsx(resistome_samdata, "../Output_files/Phyloseq_objects/Sample_data_resistome_VGO_COPD.xlsx", rownames= TRUE)


zenodo_ps@sam_data$copdcaco <- revalue(zenodo_ps@sam_data$copdcaco, c('1' ='COPD', '0' = 'control'))
zenodo_ps@sam_data$copdcaco
zenodo_ps
saveRDS(zenodo_ps,"../Output_files/Phyloseq_objects/Phyloseq_resistome_VGO_COPD.rds")
```
