---
title: "4 - Stacked bar plots and heatmaps"
author: "b.cornuhewitt@uu.nl"
---
# Packages
```{r}
library(ggplot2); library(phyloseq); library(patchwork); library(dplyr); library(viridis)
```
# Read in the ps object
```{r}
ps <- readRDS("../Output_files/Phyloseq_objects/4_COPD_resistome_phyloseq_object_rarefied_FPKM.Rds")
ps_aggl <- readRDS("../Output_files/Phyloseq_objects/5_COPD_resistome_phyloseq_object_rarefied_FPKM_90clustered.RDS")
```

# Stacked bar charts
## Absolute abundances
### AMR class
#### Individual sample bars
```{r}
# COPD cases
AMRclass_absolute_copd <- ps_aggl
# Prune ps to keep only COPD cases
AMRclass_absolute_copd <- prune_samples(sample_data(AMRclass_absolute_copd)$copdcaco == "1", AMRclass_absolute_copd)
# Make bar plot
AMRclass_absolute_bars_copd <- plot_bar(AMRclass_absolute_copd, fill = "ARG_class") +
  labs(title = "COPD Only") +
  theme(axis.text.x = element_text(angle = 270, vjust = 0.4, hjust = 0, size = 8)) 
# Save plot as svg
ggsave(filename = "../Output_files/AMR_abundance_plots/Absolute_abundances/AMRclass_absolute_bars_copd.svg", plot = AMRclass_absolute_bars_copd)


# Controls
AMRclass_absolute_control <- ps_aggl
# Prune ps to keep only controls
AMRclass_absolute_control <- prune_samples(sample_data(AMRclass_absolute_control)$copdcaco == "0", AMRclass_absolute_control)
# Create bar plot
AMRclass_absolute_bars_control <- plot_bar(AMRclass_absolute_control,
                                           fill = "ARG_class")+ labs(title = "Control Only") + theme(axis.text.x = element_text(angle = 270, vjust = 0.4, hjust = 0, size = 8))
# Save plot as svg
ggsave(filename = "../Output_files/AMR_abundance_plots/Absolute_abundances/AMRclass_absolute_bars_control.svg", plot = AMRclass_absolute_bars_control)


# Blanks
AMRclass_absolute_blanks <- ps_aggl
# Prune ps to keep only controls
AMRclass_absolute_blanks <- prune_samples(sample_data(AMRclass_absolute_blanks)$copdcaco == "blanc", AMRclass_absolute_blanks)
# Create bar plot
AMRclass_absolute_bars_blanks <- plot_bar(AMRclass_absolute_blanks, 
                                          fill = "ARG_class") + labs(title = "Blanks Only") +
  theme(axis.text.x = element_text(angle = 270, vjust = 0.4, hjust = 0, size = 8))
# Save as svg
ggsave(filename = "../Output_files/AMR_abundance_plots/Absolute_abundances/AMRclass_absolute_bars_blanks.svg", plot = AMRclass_absolute_bars_blanks)

# Combine plots using patchwork
combined_plots_absolute <- (
  AMRclass_absolute_bars_control|
  AMRclass_absolute_bars_copd|
  AMRclass_absolute_bars_blanks
)

# Set a common y-axis label
common_y_axis_label <- "Abundance"
# Add a common title for the entire layout
COPDcontrolblank_plots_absolute <- combined_plots_absolute +
  plot_layout(ncol = 3, byrow = TRUE, guides = "collect") +
  plot_annotation(title = "Total rarefied, gene length corrected and 16S qPCR corrected resistome absolute abundance", theme = theme(plot.title = element_text(hjust = 0.5, size = 20)))
# Save combined plots as svg
ggsave(filename = "../Output_files/AMR_abundance_plots/Absolute_abundances/COPDcontrolblank_plots_absolute.svg", plot = COPDcontrolblank_plots_absolute)

```
#### Grouped bars
```{r}
# Function to create grouped bar charts based on absolute abundance
bar_graphs_absolute <- function(physeq, var1 = "ungrouped", level, displayx = 10, filetype = ".jpg") {
  if (level == "Species") {
    print("Error this level is too fine for amplicon sequencing")
  } else if (var1 == "ungrouped") {
    physeqxu <- tax_glom(physeq, level)
    physeqxu <- prune_taxa(names(sort(taxa_sums(physeqxu), TRUE))[1:displayx], physeqxu)
    barchartxu <- plot_bar(physeqxu, fill = level) +
      labs(x = "Study population",
           y = "Absolute abundance",
           title = paste("Top", displayx, level, "ungrouped")) +
      scale_fill_brewer(palette = "Spectral")
    return(barchartxu)
  } else {
    physeqm <- merge_samples(physeq, var1)
    sample_data(physeqm)[[var1]] <- levels(sample_data(physeqm)[[var1]])
    physeqg <- tax_glom(physeqm, level)
    physeqg <- prune_taxa(names(sort(taxa_sums(physeqg), TRUE))[1:displayx], physeqg)
    grouped_bar <- plot_bar(physeqg, fill = level) +
      labs(x = "Study population",
           y = "Absolute abundance",
           title = paste("Top", displayx, level, sep = " ")) +
      scale_fill_brewer(palette = "Spectral")
    return(grouped_bar)
  }
}

# Agglomerate at the AMR class level
AMRclass_aggl <- tax_glom(ps, taxrank = rank_names(ps)[1]) # Rank 1 = ARG class

# COPD, control and blanks
AMRclass_absolute_bars_groupedcopdcaco <- bar_graphs_absolute(AMRclass_aggl, "copdcaco", "ARG_class") +
  ggtitle("Total rarefied, gene length corrected and 16S qPCR corrected resistome absolute abundance") +
  labs(fill = "AMR class") +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text.x = element_text(angle = 0, hjust = 0.5),  # Align labels to the center of tick points
        title = element_text(size = 17),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        legend.position = "right") +
  scale_x_discrete(labels = c("Control", "COPD", "Blank"))

ggsave(filename = "../Output_files/AMR_abundance_plots/Absolute_abundances/AMRclass_absolute_bars_groupedcopdcaco.svg", plot = AMRclass_absolute_bars_groupedcopdcaco)

# COPD and control
n_colors <- length(unique(AMRclass_absolute_bars_copd$data$ARG_class))
color_palette <- viridis(n_colors)
# Remove blanks 
AMRclass_aggl_noblanks <- AMRclass_aggl
AMRclass_aggl_noblanks <- prune_samples(!sample_data(AMRclass_aggl_noblanks)$copdcaco == "blanc", AMRclass_aggl_noblanks)
# Extract the fill colours from the previous plots to match
fill_colours <- c("#f8766d","#cd9600","#7cae00","#00be67","#00bfc4","#00a9ff","#c77cff","#ff61cc")
# Convert copdcaco to COPD and control 
AMRclass_aggl_noblanks@sam_data$copdcaco <- ifelse(AMRclass_aggl_noblanks@sam_data$copdcaco == 0, "Control", "COPD")

AMRclass_aggl_noblanks_groupedcopdcaco <- bar_graphs_absolute(AMRclass_aggl_noblanks, "copdcaco", level = "ARG_class") +
  ggtitle("Total rarefied, gene length corrected and 16S qPCR corrected resistome absolute abundance") +
  scale_fill_manual(values = fill_colours) +  # Use the extracted colors
  labs(fill = "AMR class") +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        title = element_text(size = 17),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        legend.position = "right")

ggsave(filename = "../Output_files/AMR_abundance_plots/Absolute_abundances/AMRclass_aggl_noblanks_groupedcopdcaco.svg", plot = AMRclass_aggl_noblanks_groupedcopdcaco)

```
## Relative abundances
### AMR class
#### Individual sample bars
```{r}
# Make counts relative
AMRclass_relative <- transform_sample_counts(AMRclass_aggl, function(x) x / sum(x) * 100)

# COPD
AMRclass_relative_copd<- AMRclass_relative
AMRclass_relative_copd<- prune_samples(sample_data(AMRclass_relative_copd)$copdcaco == "1", AMRclass_relative_copd)
# Make plot 
AMRclass_relative_bars_copd <- plot_bar(AMRclass_relative_copd, fill = "ARG_class") +
  labs(title = "COPD Only") +
  theme(axis.text.x = element_text(angle = 270, vjust = 0.4, hjust = 0, size = 8)) 
# Save plot
ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/AMRclass_relative_bars_copd.svg", plot = AMRclass_relative_bars_copd)

# Controls
AMRclass_relative_control <- AMRclass_relative
AMRclass_relative_control <- prune_samples(sample_data(AMRclass_relative_control)$copdcaco == "0", AMRclass_relative_control)
# Make plot 
AMRclass_relative_bars_control <- plot_bar(AMRclass_relative_control, fill = "ARG_class") +
  labs(title = "Control Only") +
  theme(axis.text.x = element_text(angle = 270, vjust = 0.4, hjust = 0, size = 8))
# Save plot
ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/AMRclass_relative_bars_control.svg", plot = AMRclass_relative_bars_control)

# Blanks
AMRclass_relative_blanks <- AMRclass_relative
AMRclass_relative_blanks <- prune_samples(sample_data(AMRclass_relative_blanks)$copdcaco == "blanc", AMRclass_relative_blanks)
# Make plot 
AMRclass_relative.bars.blanks <- plot_bar(AMRclass_relative_blanks, fill = "ARG_class") +
  labs(title = "Blanks Only") +
  theme(axis.text.x = element_text(angle = 270, vjust = 0.4, hjust = 0, size = 8))
# Save plot
ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/AMRclass_relative.bars.blanks.svg", plot = AMRclass_relative.bars.blanks)

# Combine plots using patchwork
combined_plots_relative <- (
  AMRclass_relative_bars_copd |
  AMRclass_relative_bars_control |
  AMRclass_relative.bars.blanks
)

# Set a common y-axis label
common_y_axis_label <- "Abundance"
# Add a common title for the entire layout
COPDcontrolblank_plots_relative <- combined_plots_relative +
  plot_layout(ncol = 3, byrow = TRUE, guides = "collect") +
  plot_annotation(title = "Total rarefied, gene length corrected and 16S qPCR corrected resistome relative abundance", theme = theme(plot.title = element_text(hjust = 0.5, size = 20)))
# Save as png
ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/COPDcontrolblank_plots_relative.svg", plot = COPDcontrolblank_plots_relative)

# COPD, control and blanks all in one plot
AMRclass_relative.bars <- plot_bar( AMRclass_relative, fill="ARG_class") +
    labs( title="Total rarefied, gene length corrected and 16S qPCR corrected resistome relative abundance", x="Sample", y="Abundance (%)") 
AMRclass_relative.bars + theme( axis.text.x = element_text( angle=270, vjust=0.4, hjust=0, size=8 ) )
# Save as svg
ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/AMRclass_relative.bars.svg", plot = AMRclass_relative.bars)

# COPD and controls all in one plot
blanks <- c("veldbl16", "veldbl3", "veldbl5")
AMRclass_relative_noblanks <- AMRclass_relative
AMRclass_relative_noblanks <-  prune_samples(!sample_names(AMRclass_relative) %in% blanks, AMRclass_relative)
AMRclass_relative_noblanks;AMRclass_relative 

# Create plot
AMRclass_relative_bars_noblanks <- plot_bar(AMRclass_relative_noblanks, fill="ARG_class") +
    labs( title="Total rarefied, gene length corrected and 16S qPCR corrected resistome relative abundance", x="Sample", y="Abundance (%)") 
AMRclass_relative_bars_noblanks + theme( axis.text.x = element_text( angle=270, vjust=0.4, hjust=0, size=5 ) )
# Save plot
ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/AMRclass_relative_bars_noblanks.svg", plot = AMRclass_relative_bars_noblanks)
```
#### Grouped bars
```{r}
# Function to produce grouped bar charts for relative abundance
bar.graphs.relative <- function(physeq, var1 = "ungrouped", level, displayx = 10, filetype = ".jpg") {
  if (level == "Species") {
    print("Error this level is too fine for amplicon sequencing")
  } else if (var1 == "ungrouped") {
    physeqxu <- tax_glom(physeq, level)
    # physeqxu <- transform_sample_counts(physeqxu, function(x) 100 * x/sum(x)) # removed line since I've already transformed my ps object to relative counts
    physeqxu <- prune_taxa(names(sort(taxa_sums(physeqxu), TRUE))[1:displayx], physeqxu)
    barchartxu <- plot_bar(physeqxu, fill = level) +
      labs(x = "Study population",
           y = "% of total AMR composition",
           title = paste("Top", displayx, level, "ungrouped")) +
      scale_fill_brewer(palette = "Spectral")
    return(barchartxu)
  } else {
    physeqm <- merge_samples(physeq, var1)
    sample_data(physeqm)[[var1]] <- levels(sample_data(physeqm)[[var1]])
    physeqg <- tax_glom(physeqm, level)
    physeqg <- transform_sample_counts(physeqg, function(x) 100 * x/sum(x))
    physeqg <- prune_taxa(names(sort(taxa_sums(physeqg), TRUE))[1:displayx], physeqg)
    grouped_bar <- plot_bar(physeqg, fill = level) +
      labs(x = "Study population",
           y = "% of total AMR composition",
           title = paste("Top", displayx, level, sep = " ")) +
      scale_fill_brewer(palette = "Spectral")
    return(grouped_bar)
  }
}


# New plot 
n_colors <- length(unique(AMRclass_absolute_bars_copd$data$ARG_class))
color_palette <- viridis(n_colors)

# Remove blanks from dataset
AMRclass_relative_noblanks <- AMRclass_relative
AMRclass_relative_noblanks <- prune_samples(!sample_data(AMRclass_relative_noblanks)$copdcaco == "blanc", AMRclass_relative_noblanks)

# Extract the fill colours from the previous plots to match
fill_colours <- c("#f8766d","#cd9600","#7cae00","#00be67","#00bfc4","#00a9ff","#c77cff","#ff61cc")

# Relabel copdcaco
AMRclass_relative_noblanks@sam_data$copdcaco <- ifelse(AMRclass_aggl_noblanks@sam_data$copdcaco == 0, "Control", "COPD")

AMRclass_aggl_noblanks_groupedcopdcaco <- bar.graphs.relative(AMRclass_relative_noblanks, "copdcaco", level = "ARG_class") +
  ggtitle("Total rarefied, gene length corrected and 16S qPCR corrected resistome absolute abundance") +
  scale_fill_manual(values = fill_colours) +  # Use the extracted colors
  labs(fill = "AMR class") +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text.x = element_text(angle = 0, hjust = 0.5),  # Align labels to the center of tick points
        title = element_text(size = 17),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        legend.position = "right")

ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/AMRclass_relative_bars_groupedcopdcaco.svg", plot = AMRclass_relative_bars_groupedcopdcaco)
```
### ARG cluster 
#### Individual sample bars
```{r}
# Function to calculate top 10 taxa at different levels of taxonomy
toptaxa <- function( ps, rank="ARGCluster", top=10, samples=NA, NArm=FALSE ) {
  if( ! is.na(samples[1]) ) {
    pstemp = prune_samples( samples, ps)
  } else {
    pstemp = ps
  }
  pstemp = tax_glom2( pstemp, taxrank = rank, NArm=NArm )
  pstemp = transform_sample_counts( pstemp, function(x) 100*x/sum(x) )

  if( ntaxa(pstemp) < top ) {
    top = ntaxa(pstemp)
    warning("Number of taxa in object is less then requested top list! Reduced to ", top)
  }

  toplist = names( sort( taxa_sums(pstemp), decreasing = TRUE ) )[1:top]
  return( toplist )
}

# Ps agglomerated at ARG cluster level and blanks removed = ps_aggl_noblanks
# Create new ps with only the 10 most abundant ARGs (for COPD, control & blanks) 
top10_ARGclusters_overall <- toptaxa(ps_aggl, rank="ARGCluster", top=10, samples=NA, NArm=FALSE )
top10_ARGclusters_overall
# Create new ps with only the 10 most abundant ARGs (for COPD, control & blanks) 
top10_ARGclusters_overall_noblanks<- toptaxa(ps_aggl_noblanks, rank="ARGCluster", top=10, samples=NA, NArm=FALSE )
top10_ARGclusters_overall_noblanks

# calculate Total Abundance
total_abundance <- sum(taxa_sums(ps_aggl_noblanks))
# Sum Abundances of Top-10 Taxa
top_taxa_abundance <- sum(taxa_sums(ps_aggl_noblanks)[top10_ARGclusters_overall_noblanks])
# Calculate Proportion
print(top_taxa_abundance / total_abundance)

# Top 10 in copd samples 
top10_ARGclusters_COPD<- toptaxa(COPD_ps, rank="ARGCluster", top=10, samples=NA, NArm=FALSE )
top10_ARGclusters_COPD

# Top 10 in control samples 
top10_ARGclusters_control<- toptaxa(control_ps, rank="ARGCluster", top=10, samples=NA, NArm=FALSE )
top10_ARGclusters_control

# Present as actual ARG cluster names 
tt <- data.frame(tax_table(ps_aggl_noblanks))
# extract the names of the ARGCluster column where the row names of tt match the taxa present in the top10.ARGclusters.noblanks, top10_ARGclusters_COPD and top10_ARGclusters_control
top10clusternames_noblanks <- tt$ARGCluster[ row.names(tt) %in% top10_ARGclusters_overall_noblanks] 
top10clusternames_copd <- tt$ARGCluster[ row.names(tt) %in% top10_ARGclusters_COPD]
top10clusternames_control <- tt$ARGCluster[ row.names(tt) %in% top10_ARGclusters_control]

# Create a new ps object which holds only the 10 most abundant ARGs (for all COPD and control samples)
desired_taxa_names <- top10_ARGclusters_overall_noblanks
taxa_to_keep <- taxa_names(ps_aggl_noblanks) %in% desired_taxa_names
Final_ps_top10ARGClusters <- prune_taxa(taxa_to_keep, ps_aggl_noblanks)
Final_ps_top10ARGClusters@otu_table
colnames(Final_ps_top10ARGClusters@otu_table)
top10_ARGclusters_overall_noblanks

Resistome_ps_noblanks_names <- ps_aggl_noblanks
colnames(Final_ps_top10ARGClusters@otu_table) <- top10clusternames_noblanks
rownames(Final_ps_top10ARGClusters@tax_table) <- top10clusternames_noblanks


# Make a barplot so that for COPD case and control groups, I present the relative abundance of each of the top 10 ARG classes
bar.graphs2 <- function(physeq, var1 = "ungrouped", level, displayx = 10, filetype = ".jpg") {
  if (level == "Species") {
    print("Error this level is too fine for amplicon sequencing")
  } else if (var1 == "ungrouped") {
    physeqxu <- tax_glom(physeq, level)
    physeqxu <- transform_sample_counts(physeqxu, function(x) 100 * x/sum(x))
    physeqxu <- prune_taxa(names(sort(taxa_sums(physeqxu), TRUE))[1:displayx], physeqxu)
    barchartxu <- plot_bar(physeqxu, fill = level) +
      labs(x = "COPD or control samples",
           y = "% of top 10 ARG clusters present",
           title = paste("Top", displayx, level, "ungrouped")) +
      scale_fill_brewer(palette = "Spectral")
    return(barchartxu)
  } else {
    physeqm <- merge_samples(physeq, var1)
    sample_data(physeqm)[[var1]] <- levels(sample_data(physeqm)[[var1]])
    physeqg <- tax_glom(physeqm, level)
    physeqg <- transform_sample_counts(physeqg, function(x) 100 * x/sum(x))
    physeqg <- prune_taxa(names(sort(taxa_sums(physeqg), TRUE))[1:displayx], physeqg)
    grouped_bar <- plot_bar(physeqg, fill = level) +
      labs(x = "COPD or control samples",
           y = "% of top 10 ARG clusters present",
           title = paste("Top", displayx, level, sep = " ")) +
      scale_fill_brewer(palette = "Spectral")
    return(grouped_bar)
  }
}

# Create a new phyloseq object with only the top 10 ARG clusters
ARGcluster_top10 <- prune_taxa(names(sort(taxa_sums(Final_ps_top10ARGClusters), decreasing = TRUE))[1:10], Final_ps_top10ARGClusters)

fill_colours10 <- c("#cc3311", "#4a86e8", "#ff9900", "#669966", "#33bbee", "#c77cff", "#bb5566", "#003399", "#ff33cc", "#ffcc00")

# Generate the stacked barplot at the top 10 ARG cluster level
ARGcluster_top10@sam_data$copdcaco <- revalue(ARGcluster_top10@sam_data$copdcaco, c("1"="COPD", "0"= "Control"))
suppressWarnings(library(cowplot))
theme_set(theme_cowplot())
ARGcluster_barchart <- bar.graphs2(ARGcluster_top10, "copdcaco", "ARGCluster") +
  ggtitle("Relative Abundance of Top 10 ARG Clusters in COPD and Control Groups") +
    scale_fill_manual(values = fill_colours10) +  # Use your custom color palette
  labs(fill = "ARG Cluster") +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text.x = element_text(angle = 0, hjust = 0.5),  # Align labels to the center of tick points
        title = element_text(size = 17),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        legend.position = "right") 
ARGcluster_barchart

ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/ARGcluster_barchart.svg", plot = ARGcluster_barchart)
```
# Heatmaps
## ARG level (unclustered)
```{r}
# Extract abundance matrix
ARGs <- as(otu_table(ps), "matrix")
# transform into a data frame
ARG_dataframe <-as.data.frame(ARGs)

# Transform the raw counts of reads to proportions of total ARGs in each sample 
ARG_dataframe_prop <- ARG_dataframe/rowSums(ARG_dataframe)
ARG_dataframe_prop
# Calculate the Bray-Curtis dissimilarity matrix on the full dataset
ARG_BCdist <- vegdist(ARG_dataframe_prop, method = "bray")
# Perform hierarchical cluster analysis using the BC dissimilarities 
# Each sample is assigned to its own cluster and then the algorithm proceeds iteratively, at each stage joining the two most similar clusters, continuing until there is just a single cluster.
sample_clust <- hclust(ARG_BCdist, "aver")
# make the heatmap with clustering of the samples (rows) (Rowv = as.dendrogram(sample_clust))
# Note that the samples are not in the same order as that in the matrix - heatmap() reorders both variables and observations using the BC clustering algorithm: it computes the distance between each pair of rows and columns and try to order them by similarity.
heatmap_ARGlevel_clustered <- heatmap(as.matrix(ARG_dataframe_prop), Rowv = as.dendrogram(sample_clust), Colv = NA, margins = c(10, 3))
plot(heatmap_ARGlevel_clustered)
# Add additional dendogram (on the columns) to heatmap 
# Calculate the Bray-Curtis dissimilarity matrix on ARGs in the dataset
ARG_BCdist_t <- vegdist(t(ARG_dataframe_prop), method = "bray") # We transpose the data frame so samples are now in columns
# Perform hierarchical cluster analysis using the BC dissimilarities between the ARGs (columns in dataset)
col_clust <- hclust(ARG_BCdist_t, "aver")

# create the heatmap with row and column dendograms
heatmap_ARGlevel_clustered_rowcolumn <- heatmap(as.matrix(ARG_dataframe_prop), Rowv = as.dendrogram(sample_clust), Colv = as.dendrogram(col_clust), margins = c(11, 5), xlab = "ARGs", ylab = "SampleIDs", main = "Hierarchical clustering heatmap")

# Add colour code to rows to visualise whether COPD/controls/blanks cluster together
metadata_final_ps <- sample_data(ps)
COPDcontrolblank <- metadata_final_ps$copdcaco 
COPDcontrolblank<- replace(COPDcontrolblank, which(COPDcontrolblank == "0"), "blue")
COPDcontrolblank<-replace(COPDcontrolblank, which(COPDcontrolblank == "1"), "red")
COPDcontrolblank<-replace(COPDcontrolblank, which(COPDcontrolblank == "blanc"),"green")

# Add these colour codes to the heatmap 
heatmap_ARGlevel_clustered_colour1 <- heatmap(as.matrix(ARG_dataframe_prop), Rowv = as.dendrogram(sample_clust), Colv = as.dendrogram(col_clust), RowSideColors = COPDcontrolblank, margins = c(11, 5), xlab = "ARGs", ylab = "SampleIDs", main = "Hierarchical clustering heatmap")
```
## ARG cluster level 
```{r}
# Create a heatmap now on the cluster 90 level as this is a better representation of the data 
# Extract abundance matrix from cluster 90 agglomerated ps object
ARGclustertable <- as(otu_table(ps_aggl), "matrix")
# transform into a data frame
ARGcluster_dataframe <- as.data.frame(ARGclustertable)
ARGcluster_dataframe
# Examine actual cluster 90 names and names assigned in the ps object
ps_aggl@tax_table[,6]
# column names are (specific) ARG names, agglomeration is correct (lowest level is ARG cluster) but col names are not correct - I will rename the column names to the AMR classes for presentation in the heatmap
colnames(ARGcluster_dataframe) <- c("aac(3)-II_clust","aac(3)-Iva","aac(6')-aph(2'')_clust","aadA_ant(3'')-Ia_clust","aadA_clust1","aadD","ant(6)-Ia_clust2","aph(3'')-Ib","aph(3')-Ia_aph(3')-Ic","aph(3')-Ib","aph(3')-III","aph(4)-Ia","aph(6)-Id","str","blaACT_clust","blaACT_CMG_MIR_clust","blaBRO","blaCARB_clust2","blaCTX-M_clust1","blaOXA-22","blaOXA-395_clust","blaOXA-60_clust","blaOXA-85","blaOXA_clust19","blaOXA_clust3","blaOXA_clust8","blaOXA_clust9","blaOXY_clust1","blaSPU-1","blaTEM_clust","blaZ_clust","cfxA_clust","mecA_clust","mecA1","penA","fusB","lsa(A)","lsa(C)","erm(A)_2_AF002716","erm(B)_clust","erm(C)_clust","erm(F)_clust","erm(T)_4_AJ488494","erm(X)_clust","lnu(C)","mdf(A)","mef(A)-3","mef(A)_clust","mph(A)","mph(C)","msr(A)","msr(D)","msr(E)","vga(A)_clust","cat(pC194)","cat_2","catA1","catQ","catS","cml_clust","cmx","cfr(C)","sul1","sul2","tet(32)","tet(33)","tet(37)","tet(39)","tet(40)","tet(44)","tet(A)","tet(B)","tet(C)","tet(G)","tet(K)","tet(L)_clust1","tet(M)","tet(O)","tet(O/32/O)","tet(O/W/32/O/W/O)","tet(O/W/O)-1","tet(Q)","tet(T)","tet(W)","dfrA15_clust")
# Check the renaming has been performed correctly
colnames(ARGcluster_dataframe)
# transform the raw counts of reads to proportions of total ARG clusters in each sample 
ARGcluster_dataframe_prop <- ARGcluster_dataframe/rowSums(ARGcluster_dataframe)
# Examine the dataframe 
ARGcluster_dataframe_prop
# calculate the Bray-Curtis dissimilarity matrix on the full dataset
ARGCluster_BCdist <- vegdist(ARGcluster_dataframe_prop, method = "bray")
# perform hierarchical cluster analysis using the BC dissimilarities 
# each sample is assigned to its own cluster and then the algorithm proceeds iteratively, at each stage joining the two most similar clusters, continuing until there is just a single cluster.
ARGCluster_clust <- hclust(ARGCluster_BCdist, "aver")
# Create a heatmap at the ARG cluster level
heatmap_ARGClusterlevel <- heatmap(as.matrix(ARGcluster_dataframe_prop), Rowv = as.dendrogram(ARGCluster_clust), RowSideColors = COPDcontrolblank, margins = c(11, 5), xlab = "ARGCluster", ylab = "SampleIDs", main = "Hierarchical clustering heatmap at 90% ARG cluster level")
# Note that the samples are not in the same order as that in the matrix - heatmap() reorders both variables and observations using the BC clustering algorithm: it computes the distance between each pair of rows and columns and try to order them by similarity.


# Using tutorial: https:/slowkow.com/notes/pheatmap-tutorial/
# Use the data frame which I created previously for ARG cluster level 
ARGcluster_dataframe_prop
# convert ARG Cluster df to matrix form - this is needed for pheatmap function
ARGcluster_prop_mat <- as.matrix(ARGcluster_dataframe_prop)
# check the matrix to see it is formulated correctly
rownames(ARGcluster_prop_mat)
colnames(ARGcluster_prop_mat)
# create a new df with column for copdcaco status - in order to label the heatmap by sample type (colour coding later on)
ARGcluster_df_copdcaco <- ARGcluster_dataframe_prop
ARGcluster_df_copdcaco$copdcaco <- ps@sam_data$copdcaco
copdcaco_df <- data.frame("Population" = ARGcluster_df_copdcaco$copdcaco)

copdcaco_df$Population <- ifelse(copdcaco_df$Population == '1', 'COPD',
                               ifelse(copdcaco_df$Population == '0', 'Control',
                                      ifelse(copdcaco_df$Population == 'blanc', 'Blank', copdcaco_df$Population
)))
rownames(copdcaco_df) = rownames(ARGcluster_df_copdcaco) # name matching
# specify the colours wanted as the key (annotation_colours argument in pheatmap function) - red-COPD, blue-control, green-blank

row_colors = list(Population = c("COPD" = "orange", "Control" = "cornflowerblue", "Blank"= "green3"))

# Specify BC clustering for pheatmap 
BCdist_rows <- vegdist(ARGcluster_dataframe_prop, method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE)
BCdist_cols <- vegdist(t(ARGcluster_dataframe_prop), method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE)
pheatmap(ARGcluster_prop_mat, cluster_rows=FALSE, cluster_cols=FALSE)
# Create heatmap including all ARG clusters. BC distance clustering, colour coding for COPD/control/blank samples on y-axis
ARG_cluster_heatmap <- pheatmap(
  mat = ARGcluster_prop_mat,
  color = colorRampPalette(rev(brewer.pal(n = 10, name = "RdYlBu")))(100), # see other colour palettes: https:/www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette/ (also can change from (10) to another number to change density of colour), this one is colour blind friendly
  border_color = "grey60",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_distance_rows = BCdist_rows, # using the BC values calculated previously 
  clustering_distance_cols = BCdist_cols ,
  show_colnames = TRUE,
  show_rownames = TRUE,
  annotation_row = copdcaco_df, # annotates COPD/control/blanks and colour codes them - I want to change the colours though to red COPD, blue control, green blank
  annotation_colors = row_colors, # change colours to orange COPD, blue control, green blank 
  drop_levels = TRUE,
  fontsize= 15,
  fontsize_row = 10,
  fontsize_col= 10,
  angle_col = 45,
  main= "Heatmap at ARG cluster level", 
  legend_title = "Sample")

ggsave("../Output_files/Heatmaps/ARG_cluster_heatmap.svg", ARG_cluster_heatmap, width = 18, height = 12)
```
## Heatmap at AMR class level 
```{r}
# Create a heatmap on the AMR class level
AMRclasstable <- as(otu_table(AMRclass_aggl), "matrix")
# transform into a data frame
AMRclass_dataframe <- as.data.frame(AMRclasstable)
AMRclass_dataframe
# column names are (specific) ARG names, agglomeration is correct (lowest level is AMR class) but col names are not correct - I will rename the column names to the AMR classes for presentation in the heatmap
colnames(AMRclass_dataframe) <- c("Aminoglycoside", "Beta-Lactam", "Macrolide", "Tetracycline", "Phenicol", "Sulphonamide", "Trimethoprim", "Fusidicacid")
# Check the renaming has been performed correctly
AMRclass_dataframe
view(AMRclass_dataframe)
# transform the raw counts of reads to proportions of total AMR class in each sample 
AMRclass_dataframe_prop <- AMRclass_dataframe/rowSums(AMRclass_dataframe)
# examine the dataframe 
AMRclass_dataframe_prop
AMRclass_dataframe_prop[1:3, 1:3]
# calculate the Bray-Curtis dissimilarity matrix on the full dataset
AMRclass_BCdist <- vegdist(AMRclass_dataframe_prop, method = "bray")
# perform hierarchical cluster analysis using the BC dissimilarities 
# each sample is assigned to its own cluster and then the algorithm proceeds iteratively, at each stage joining the two most similar clusters, continuing until there is just a single cluster.
sample_clust2 <- hclust(AMRclass_BCdist, "aver")
# perform hierarchical cluster analysis using the BC dissimilarities between the AMR classes (columns in dataset)
AMRclass_clust <- hclust(AMRclass_BCdist, "aver")
# Create a heatmap at the AMR class level
heatmap_AMRclasslevel <- heatmap(as.matrix(AMRclass_dataframe_prop), Rowv = as.dendrogram(sample_clust2), RowSideColors = COPDcontrolblank, margins = c(11, 5), xlab = "AMR class", ylab = "SampleIDs", main = "Hierarchical clustering heatmap at AMR class level")
# Note that the samples are not in the same order as that in the matrix - heatmap() reorders both variables and observations using the BC clustering algorithm: it computes the distance between each pair of rows and columns and try to order them by similarity.


# Heatmap at AMR class level (using pheatmap)
AMRclass_dataframe_prop # use previous df created
# convert AMR Class df to matrix form - this is needed for pheatmap function
AMRclass_prop_mat <- as.matrix(AMRclass_dataframe_prop)
# check the matrix to see it is formulated correctly
rownames(AMRclass_prop_mat)
colnames(AMRclass_prop_mat)
# create a new df with column for copdcaco status - in order to label the heatmap by sample type (colour coding later on)
AMRclass_dataframe_prop_copdcaco <- AMRclass_dataframe_prop
AMRclass_dataframe_prop_copdcaco$copdcaco <- ps@sam_data$copdcaco
AMRclass_copdcaco_df<- data.frame("COPD/control/blank" = AMRclass_dataframe_prop_copdcaco$copdcaco)
rownames(AMRclass_copdcaco_df) = rownames(AMRclass_dataframe_prop_copdcaco) # name matching
# specify the colours wanted as the key (annotation_colours argument in pheatmap function) - red-COPD, blue-control, green-blank
row_colors_AMRclass = list( COPD_control_blank = c("1" = "#FF0000", "0" = "#0000FF", "blanc"= "#00FF00" ))
# Specify BC clustering for pheatmap - default is euclidean(?)
AMRclass_BCdist_rows <- vegdist(AMRclass_dataframe_prop, method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE)
AMRclass_BCdist_cols <- vegdist(t(AMRclass_dataframe_prop), method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE)
# Create heatmap including all ARG clusters. BC distance clustering, colour coding for COPD/control/blank samples on y-axis
pheatmap(
  mat = AMRclass_prop_mat,
  color = colorRampPalette(rev(brewer.pal(n = 10, name = "RdYlBu")))(100), # see other colour palettes: https:/www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette/ (also can change from (10) to another number to change density of colour)
  border_color = "grey60",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_distance_rows = AMRclass_BCdist_rows, # using the BC values calculated previously 
  clustering_distance_cols = AMRclass_BCdist_cols ,
  show_colnames = TRUE,
  show_rownames = TRUE,
  annotation_row = AMRclass_copdcaco_df, # annotates COPD/control/blanks and colour codes them - I want to change the colours though to red COPD, blue control, green blank
  annotation_colors = row_colors_AMRclass, # change colours to red COPD, blue control, green blank (see annotat) 
  drop_levels = TRUE,
  fontsize= 20,
  fontsize_row = 10,
  fontsize_col= 20,
  main= "Heatmap at AMR class level"
)
```
# Plot for manuscript
```{r}
# Stacked bar plots
bar.graphs.relative <- function(physeq, var1 = "ungrouped", level, displayx = 10, filetype = ".jpg") {
  if (level == "Species") {
    print("Error this level is too fine for amplicon sequencing")
  } else if (var1 == "ungrouped") {
    physeqxu <- tax_glom(physeq, level)
    # physeqxu <- transform_sample_counts(physeqxu, function(x) 100 * x/sum(x)) # removed line since I've already transformed my ps object to relative counts
    physeqxu <- prune_taxa(names(sort(taxa_sums(physeqxu), TRUE))[1:displayx], physeqxu)
    barchartxu <- plot_bar(physeqxu, fill = level) +
      labs(x = "Study population",
           y = "% of total AMR composition",
           title = paste("Top", displayx, level, "ungrouped")) +
      scale_fill_brewer(palette = "Spectral")
    return(barchartxu)
  } else {
    physeqm <- merge_samples(physeq, var1)
    sample_data(physeqm)[[var1]] <- levels(sample_data(physeqm)[[var1]])
    physeqg <- tax_glom(physeqm, level)
    physeqg <- transform_sample_counts(physeqg, function(x) 100 * x/sum(x))
    physeqg <- prune_taxa(names(sort(taxa_sums(physeqg), TRUE))[1:displayx], physeqg)
    grouped_bar <- plot_bar(physeqg, fill = level) +
      labs(x = "Study population",
           y = "% of total AMR composition",
           title = paste("Top", displayx, level, sep = " ")) +
      scale_fill_brewer(palette = "Spectral")
    return(grouped_bar)
  }
}

# AMR class 
AMRclass_aggl_noblanks_groupedcopdcaco <- bar.graphs.relative(AMRclass_relative_noblanks, "copdcaco", level = "ARG_class")+
  scale_fill_manual(values = fill_colours) +  # Use the extracted colors
  labs(fill = "AMR class", title = NULL) +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text.x = element_text(angle = 0, hjust = 0.5),  
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        legend.position = "right")

AMRclass_aggl_noblanks_groupedcopdcaco

# ARG cluster level 
# Prune ps object to contain only the top 10 ARG clusters
ARGcluster_top10 <- prune_taxa(names(sort(taxa_sums(Final_ps_top10ARGClusters), decreasing = TRUE))[1:10], Final_ps_top10ARGClusters)

fill_colours10 <- c("#cc3311", "#4a86e8", "#ff9900", "#669966", "#33bbee", "#c77cff", "#bb5566", "#003399", "#ff33cc", "#ffcc00")

ARGcluster_top10@sam_data$copdcaco <- revalue(ARGcluster_top10@sam_data$copdcaco, c("1"="COPD", "0"= "Control"))

theme_set(theme_cowplot())
ARG_top10_relative_bar <- bar.graphs.relative(ARGcluster_top10, "copdcaco", level = "ARGCluster")+
  scale_fill_manual(values = fill_colours10) +  # Use the extracted colors
  labs(fill = "ARG Cluster", title = NULL, y = "% of top 10 ARG clusters present")+
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text.x = element_text(angle = 0, hjust = 0.5),  
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        legend.position = "right")

# Patchwork to join the plots together
patched_barcharts <- (AMRclass_aggl_noblanks_groupedcopdcaco|ARG_top10_relative_bar)
patched_bar_labelled <- patched_barcharts + plot_annotation(tag_levels = 'a')

ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/patched_bar_labelled.svg", width = 14, height = 6, plot = patched_bar_labelled)
```


# Manuscript plots
# Compute top10 taxa
```{r}
top10<- toptaxa(ps, rank="ARGCluster", top=10, samples=NA, NArm=FALSE )
top10
```
# Figure 2 for manuscript
```{r}
#' A convenience wrapper to generate tuned barplots at various taxonomic levels using phyloseq input
#'
#' @param ps phyloseq object
#' @param taxrank taxonomic rank to show (default "Phylum")
#' @param top top-x to show (default top=10)
#' @param sort.stack whether or not to sort the actual stacking on overall abundance (default sort.stack=TRUE)
#' @param x what to show on x-axis (default x="Sample")
#' @param y what to show on y-axis (default y="Abundance")
#' @param group metadata grouping variable to "merge_samples" phyloseq function
#' @param group_fun default "mean". phyloseq merge_samples will SUM otu_table by default. Here we corrected for the Nsamples per group -> actual MEAN. All other values will result in default behaviour.
#' @param logy transform Abundance y-scale to log10 (default FALSE)
#' @param facet_grid we want to facet the stacked bars (default facet_grid=NULL, example facet_grid="Kingdom")
#' @param NArm whether or not to remove NAs in the data while agglomerating at given taxonomic rank (default NArm=FALSE)
#' @param legend.col How many columns the legend may contain (default legend.col=1)
#' @param xlab.rel Tweaked size of x-axis labels since readability depends on N samples (default xlab.rel=0.5)
#' @param title Any custom to plot on the figure (default generates "Barplot <taxrank> - top<x> taxa")
#' @param title.center center the title on figure otherwise left adjusted (default=TRUE)
#'
#' @returns ggplot figure
#'
#' @author Alex Bossers \email{a.bossers@uu.nl}
#' #' Code modified by Beatrice Cornu Hewitt on 19th March 0
#'
#' @import RColorBrewer
#' @import pals
#' @import tax_glom2
#' @import toptaxa
#'  
#' Updates:
#'          2022-07-01 alx added taxglom options to allow visible subtaxonomic ranks per bar.
#'          2022-05-15 alx added grouping and MEAN grouping options
#'          2021-xx-xx alx many small tweaks. See git history.
#'          2020-03-02 Alx split bucket into functions
#'          2020-01-01 Alx initial checkin
#' 


# FixMe: sorting abundance of stacks when not using taxglom showing subtaxonomic ranks. Sorting is different.


ps_plot_bar <- function( ps, 
                         taxrank="Phylum", top=10, taxglom=TRUE, taxglom.next="",
                         sort.stack=TRUE, 
                         x="Sample", y="Abundance", group="", group_fun="mean",
                         logy=FALSE, 
                         facet_grid=NULL, NArm=FALSE, 
                         xlab.rel = 0.5, 
                         legend=TRUE, legend.col=1, legend.size=8,
                         title=paste0("Barplot ", taxrank, " - ", if(top>0){paste0("Top",top)}else{"all taxa"} ), title.center=TRUE,
                         colpalette=NULL  # argument for specifying color palette
                       ) {
  
  require(RColorBrewer)
  require(pals)
  require(dplyr)
  require(phyloseq)

  # other custom functions from my toolbox
  #source("tax_glom2.R")
  #source("toptaxa.R")

  # first drop zero samplesum samples (throws error otherwise)
  zerosamples <- sum( sample_sums(ps) == 0 )
  if( zerosamples > 0 ) {
    warning( paste0( "Samples with zero sample_sums detected (N=", zerosamples, "). These are removed before plotting!\n" ) )
    # note subset_samples doesn't work in user functions! See Rplayground for working example demonstrating this.
    ps <- prune_samples( sample_sums(ps) > 0, ps )  
   }
  
  # Group samples if grouping variable given
  if(group > "") {
    
    # first prevent NAs in metadata variable
    if( sum( is.na( sample_data(ps)[,group] ) ) > 0 ) {
      stop("Metadata variable column contains NAs which may cause strange results. Aborting!")
    }
    ps.tmp <- merge_samples( ps, group=group )
    if( group_fun == "mean" ) {
      # otu table goes default to SUM for otu table. I want to show actual MEAN so correct for sample numbers
      snumbers <- sample_data(ps) %>% group_by_at( group ) %>% count()
      #double check if sample order is the same as metadata count order.
      ps.o <- otu_table(ps.tmp)
      if( ! all(snumbers[,1] == row.names(ps.o) ) )  {
        stop("Internal error occurred. Sample name orders doesn't seem to match order of samples_count per group somehow... Turn off group_fun='mean'.")
      } else {
        ps.o <- ps.o / snumbers$n
        otu_table(ps.tmp) <- ps.o
      }
    } else if( group_fun == "median" ) {
      # otu table goes default to SUM for otu table. I want to show actual MEAN so correct for sample numbers
      snumbers <- sample_data(ps) %>% group_by_at( group ) %>% count()
      stop("Median function not implemented yet!")
    } else {
      warning("Grouping function defaulting to SUM!")
    }
    ps <- ps.tmp
  }
  
  #glom at correct taxon level
  if(!taxglom) { 
      #keep in case taxglom=FALSE
      ps.org <- ps 
    } else { 
      ps.org <- ps 
      ps <- tax_glom2( ps, taxrank=taxrank, NArm=NArm )
    }
  
  if(top>0) {
    # make first top x at taxrank
    topx <- toptaxa( ps, rank=taxrank, top=top )
    ps <- prune_taxa( topx, ps)
  }
  
  if(!taxglom) {
    # get taxa list op taxrank level to extract later if taxglom = FALSE. 
    # Using taxa as is would return only a subset of representative ASVs so we need to select taxa names at higher level first to know exact ASVs.
    taxtab <- data.frame(tax_table(ps))
    taxtop <- taxtab[topx,]
    topPerRank <- taxtop[,taxrank] #taxa at i.e. Family level
    
    #Now we determine whether we want to taxglom deeper level or not
    #Else we leave data at ASV level.
    ps.org2 <- prune_taxa( data.frame(tax_table(ps.org))[,taxrank] %in% topPerRank, ps.org )#subset_taxa( ps.org, eval(as.name(taxrank)) %in% topPerRank)
    if(taxglom.next != "") {
      ps.org2 <- tax_glom2( ps.org2, taxrank=taxglom.next )
    }
      ps <- ps.org2
  }
  
  p <- plot_bar( ps, x=x, y=y, fill=taxrank, title=title, facet_grid=facet_grid )

  # sort stacked bars
  if(sort.stack!="FALSE") {
    if(sort.stack=="reverse") {
      warning("reverse sort")
      p.sort <- p$data %>% group_by( eval(as.name(taxrank)) ) %>% summarise( Avg=mean(Abundance)) %>% arrange( desc(Avg) )
    } else {
      p.sort <- p$data %>% group_by( eval(as.name(taxrank)) ) %>% summarise( Avg=mean(Abundance)) %>% arrange( Avg )
    }
      colnames(p.sort) <- c(taxrank,"Avg")
      p.sort
      p$data[taxrank] <- factor( p$data[[taxrank]], levels=p.sort[[taxrank]] )
  }
  
  # Use the specified color palette if provided
  if (!is.null(colpalette)) {
    p <- p + scale_fill_manual(values = colpalette )
  }
  
  p <- p + theme( text = element_text(size=10),
           axis.text.y = element_text( size=8 ) , 
           axis.text.x = element_text( angle=270, vjust=0.1, hjust=0, size=8 ),
           legend.title = element_text( size=legend.size, face="bold"),
           legend.text = element_text( size=legend.size )) +
    scale_y_continuous(expand = c(0, 0))
  
  if(legend.col) {
    p <- p + guides( fill=guide_legend(ncol=legend.col) )
  }
  
  if( ! legend ){
    p <- p + guides( fill="none" )
  }
  
  if(!taxglom) {
    #add subtitle when different taxglom is used
    if(taxglom.next == ""){
      sub <- paste0("(ASV-level taxa stacked per ",taxrank,"-taxonomic rank)")
    } else {
      sub <- paste0("(",taxglom.next,"-level taxa stacked per ",taxrank,"-taxonomic rank)")
    }
    p <- p + labs( subtitle=sub )
  }
  
  if(title.center) {
    p <- p + theme( plot.title = element_text( hjust = 0.5 ),
                    plot.subtitle = element_text( hjust = 0.5 ))
  }
  
  p <- p + theme( axis.text.x = element_text( size=rel(xlab.rel) ) )
  
  # if logy then transform y axis and label
  if(logy){
    p <- p + scale_y_continuous(trans='log10') + ylab("Log10(Abundance)")
  }

  return(p)
}

```
# Absolute abundance plots
```{r}
# ARG cluster level (top 10)
barplot_abs_ARGcluster <- ps_plot_bar( ps_noblanks, 
                        taxrank="ARGCluster", 
                        top=10, 
                        taxglom=TRUE, 
                        taxglom.next="",
                        sort.stack="reverse", 
                        x="Sample", 
                        y="Abundance", 
                        group="copdcaco", 
                        group_fun="mean",
                        logy=FALSE, 
                        facet_grid=NULL, 
                        NArm=FALSE, 
                        xlab.rel = 0.5,   
                        legend=TRUE, 
                        legend.col=1, 
                        legend.size=8,
                        colpalette = colpalette1, 
                        title = NULL)+
  scale_x_discrete(labels = c("0" = "Control", "1" = "COPD"))+
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 16),
        axis.text.y = element_text(size = 16),
        legend.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 16),
        plot.subtitle = element_text(size = 16))+
  ylab("Abundance (FPKM) (of top 10 clusters)") + 
  labs(fill = "Top 10 ARG clusters")  # Set legend title manually

              
# AMR class level
barplot_abs_AMRclass <- ps_plot_bar( Final.ps.noblanks, 
                        taxrank="ARG_class", 
                        top=10, 
                        taxglom=TRUE, 
                        taxglom.next="",
                        sort.stack="reverse", 
                        x="Sample", 
                        y="Abundance", 
                        group="copdcaco", 
                        group_fun="mean",
                        logy=FALSE, 
                        facet_grid=NULL, 
                        NArm=FALSE, 
                        xlab.rel = 0.5,   
                        legend=TRUE, 
                        legend.col=1, 
                        legend.size=8, 
                        colpalette = colpalette2, 
                        title = NULL)+
  scale_x_discrete(labels = c("0" = "Control", "1" = "COPD"))+
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 16),
        axis.text.y = element_text(size = 16),
        legend.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 16),
        plot.subtitle = element_text(size = 16))+
  ylab("Abundance (FPKM)")+
    labs(fill = "AMR class") 

# Save plots
ggsave(filename = "../Output_files/AMR_abundance_plots/Absolute_abundances/barplot_abs_ARGcluster.svg", plot = barplot_abs_ARGcluster)
ggsave(filename = "../Output_files/AMR_abundance_plots/Absolute_abundances/barplot_abs_AMRclass.svg", plot = barplot_abs_AMRclass)

# Combine plots for supplement figure
abs_barplots <- (barplot_abs_ARGcluster|barplot_abs_AMRclass)
patched_abs <- abs_barplots + plot_annotation(tag_levels = 'a')

ggsave(filename = "../Output_files/AMR_abundance_plots/Absolute_abundances/patched_abs.svg", width = 14, height = 6, plot = patched_abs)
```
# Relative abundance 
```{r}
# Convert to relative abundances
Final_ps_noblanks_rel <- transform_sample_counts(ps_aggl, function(x) x / sum(x) * 100)

# Define breaks and labels for the y-axis
y_breaks <- c(25, 50, 75, 100)
y_labels <- paste0(y_breaks, "%")
# Apply cowplot theme
theme_set(theme_cowplot())

# ARG cluster level (top10)
barplot_RA_ARGcluster <- ps_plot_bar(Final_ps_noblanks_rel, 
                                     taxrank="ARGCluster", 
                                     top=10, 
                                     taxglom=TRUE, 
                                     taxglom.next="",
                                     x="Sample", 
                                     y="Abundance", 
                                     group="copdcaco", 
                                     group_fun = "mean",
                                     logy=FALSE, 
                                     facet_grid=NULL, 
                                     NArm=FALSE, 
                                     xlab.rel = 0.5,   
                                     legend=TRUE, 
                                     legend.col=1, 
                                     legend.size=8,
                                     colpalette = colpalette1, 
                                     title = NULL) + 
  scale_x_discrete(labels = c("0" = "Control", "1" = "COPD")) +
  scale_y_continuous(breaks = y_breaks, labels = y_labels) +  # Set breaks and labels for y-axis
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 16),
        axis.text.y = element_text(size = 16),
        legend.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 16),
        plot.subtitle = element_text(size = 16)) +
  ylab("% of top 10 ARG clusters present")+
  labs(fill = "Top 10 ARG clusters")  # Set legend title manually
  
# AMR class level
barplot_RA_ARGclass <- ps_plot_bar(Final.ps.noblanks.rel, 
                                   taxrank="ARG_class", 
                                   top=10, 
                                   taxglom=TRUE, 
                                   taxglom.next="",
                                   x="Sample", 
                                   y="Abundance", 
                                   group="copdcaco", 
                                   group_fun="mean",
                                   logy=FALSE, 
                                   facet_grid=NULL, 
                                   NArm=FALSE, 
                                   xlab.rel = 0.5,   
                                   legend=TRUE, 
                                   legend.col=1, 
                                   legend.size=8,
                                   colpalette = colpalette2,
                                   title = NULL) + 
  scale_x_discrete(labels = c("0" = "Control", "1" = "COPD")) +
  scale_y_continuous(breaks = y_breaks, labels = y_labels) +  # Set breaks and labels for y-axis
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 16),
        axis.text.y = element_text(size = 16),
        legend.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 16),
        plot.subtitle = element_text(size = 16)) +
  ylab("% of total ARGs present")+
  labs(fill = "AMR class")  # Set legend title manually

# Save plots
ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/barplot_RA_ARGcluster.svg", plot = barplot_RA_ARGcluster, width = 8, height= 6)
ggsave(filename = "../Output_files/AMR_abundance_plots/Relative_abundances/barplot_RA_ARGclass.svg", plot = barplot_RA_ARGclass, width = 8, height= 6)

# Patchwork for manuscript figure
# Combine the plots
RA_barplots <- (barplot_RA_ARGclass | barplot_RA_ARGcluster)
RA_barplots <- RA_barplots+ plot_annotation(tag_levels = 'a') & 
  theme(plot.tag = element_text(size = 16))

# Save the patched figure
ggsave(
  filename = "../Output_files/AMR_abundance_plots/Relative_abundances/patched.svg", 
  width = 14, 
  height = 6, 
  plot = RA_barplots
)

```
